---
title: "Machine Learning Prediction of Prolonged ICU Stay in Sepsis Patients Using MIMIC-IV"
subtitle: "Biostat 203B Final Project"
author: 
  - name: "Charlotte Lee and 206782165"
  - name: "Yasmine Tani and Brandon Brushwyler"
date: "March 20, 2026"
execute:
  cache: true
format:
  html:
    theme: cosmo
    embed-resources: true
    number-sections: true
    toc: true
    toc-depth: 4
    toc-location: left
    code-fold: false
---

# Abstract

Sepsis is a life-threatening condition and a major cause of Intensive Care Unit (ICU) admission and mortality worldwide. Patients with sepsis who experience prolonged ICU stays (≥14 days) face worse outcomes and increased healthcare costs, yet few studies have developed prediction tools specifically targeting extended length of stay at the time of ICU admission.

This study evaluates six machine learning models for predicting prolonged ICU stay in adult sepsis patients using retrospective data from the MIMIC-IV Clinical Database (n = 4,179). Predictors included first 24-hour worst-case vital signs, laboratory values, demographics, comorbidities derived from ICD-9 and ICD-10 codes, and three clinically motivated interaction terms (CKD×Diabetes, Chronic Liver×CVD, GCS×Vasopressor). Missing values were handled through mean imputation for symmetric variables and median imputation for skewed variables. Models compared included logistic regression, LASSO, Ridge, Elastic Net, Random Forest, and XGBoost, evaluated using AUC and Youden-optimal thresholds on a held-out 20% test set.

XGBoost achieved the highest AUC (0.683), closely followed by Ridge (0.681) and LASSO (0.680). Random Forest performed comparably (AUC = 0.677). Differences across models were minimal (range 0.677– 0.683), suggesting predominantly linear relationships between predictors and prolonged stay. Key predictors included chronic liver disease, cardiovascular disease, Glasgow Coma Scale score, and insurance type. LASSO retained the CKD×Diabetes interaction while dropping diabetes as a main effect, consistent with diabetic nephropathy as the clinically meaningful entity.

These findings suggest that comorbidity burden at admission drives prolonged ICU stay more than acute physiological derangements alone, and that simpler interpretable models perform comparably to complex nonlinear approaches. Given its comparable performance and interpretable coefficients, LASSO was selected for clinical deployment in the accompanying prediction application.

# Introduction

## Background

## Literature Review

## Problem Statement

## Objectives

# Methods

## Data Source

Data were obtained from MIMIC-IV (Medical Information Mart for Intensive Care), a large single-center critical care database containing de-identified electronic health records from Beth Israel Deaconess Medical Center (BIDMC) in Boston, MA. MIMIC-IV contains data from over 50,000 ICU admissions and includes diagnoses, laboratory results, vital signs, medications, and administrative data. Access to MIMIC-IV requires completion of a data use agreement and CITI human subjects training through PhysioNet, ensuring appropriate credentialing and privacy protections. Data were accessed via Google BigQuery using the `bigrquery` and `dbplyr` R packages, enabling SQL-based querying of large tables without loading full datasets into memory. All filtering and aggregation were performed server-side on BigQuery before collecting results, consistent with best practices for handling large-scale clinical databases.

**Inclusion criteria:** Adult patients (age ≥18) with a primary sepsis diagnosis (ICD-9 or ICD-10, sequence number = 1) admitted to the ICU. Only the first qualifying admission per patient was retained to satisfy the independence assumption of logistic regression.

**Exclusion criteria.** Patients who were flagged to have died within 48 hours of ICU admission were excluded to address competing risk bias, these patients cannot by definition have a prolonged stay, and their inclusion would cause high-severity markers to appear falsely protective.

Display machine information:

```{r}
sessionInfo()
```

Display my machine memory.

```{r}
memuse::Sys.meminfo()
```

Load database libraries and the tidyverse frontend:

```{r}
#| label: setup
#| cache: false
#| message: false
#| warning: false
library(bigrquery)
library(dbplyr)
library(DBI)
library(gt)
library(gtsummary)
library(tidyverse)
library(glmnet)
library(pROC)
library(corrplot)
library(tidymodels)
library(ranger)
library(vip)
library(xgboost)
library(styler)

# Style my current file
#styler::style_file("final_project/final.qmd")
```

## Connect to The Data

```{r}
#| cache: false  # overrides the YAML cache: true for this chunk only

bq_auth(path = "biostat-203b-2026-winter-92fefbfab477.json")
con_bq <- dbConnect(
  bigrquery::bigquery(),
  project = "biostat-203b-2025-winter",
  dataset = "mimiciv_3_1",
  billing = "biostat-203b-2025-winter"
)
```

### Cohort Definition

```{r}
#| label: cohort-definition

patients_tbl <- tbl(con_bq, "patients") |>
  select(subject_id, anchor_age, gender)

sepsis_codes_tbl <- tbl(con_bq, "d_icd_diagnoses") |>
  filter(sql("LOWER(long_title) LIKE '%sepsis%'")) |>
  select(icd_code, icd_version, long_title)

sepsis_cohort_tbl <- tbl(con_bq, "diagnoses_icd") |>
  filter(seq_num == 1) |>
    inner_join(sepsis_codes_tbl, by = c("icd_code", "icd_version")) |>
    select(subject_id, hadm_id, seq_num, icd_code, icd_version, long_title)

cohort_df <-sepsis_cohort_tbl |>
inner_join(patients_tbl |> filter(anchor_age >= 18),
by = "subject_id") |>
collect()

icu_hadm_ids <- tbl(con_bq, "icustays") |>
select(hadm_id) |> distinct() |> collect() |> pull(hadm_id)

cohort_df <- cohort_df |> filter(hadm_id %in% icu_hadm_ids)

cat("Total admissions:", nrow(cohort_df), "\n")
cat("Unique patients: ", n_distinct(cohort_df$subject_id), "\n")

```

**Outcome:** Prolonged ICU stay, defined as hospital length of stay ≥14 days (binary: 1 = prolonged, 0 = normal). Length of stay was computed as the difference in days between hospital admissiona nd discharge timestamps.

```{r}
#| label: outcome-variable
admissions_tbl <- tbl(con_bq, "admissions") |>
select(subject_id, hadm_id, admittime, dischtime, race, insurance) |>
mutate(los_days = sql("TIMESTAMP_DIFF(dischtime, admittime, DAY)")) |>
collect()

sepsis_with_outcome <- cohort_df |>
inner_join(admissions_tbl, by = c("subject_id", "hadm_id")) |>
filter(!is.na(los_days), los_days > 0) |>
mutate(prolonged_stay = if_else(los_days >= 14, 1L, 0L))

sepsis_with_demo <- sepsis_with_outcome
cat("Final cohort:          ", nrow(sepsis_with_demo), "\n")

cat("Prolonged stay (>=14d):", sum(sepsis_with_demo$prolonged_stay),
"(", round(mean(sepsis_with_demo$prolonged_stay)*100,1), "%)\n")
```

## Variables and Features

**Predictors (19 total):**

**Vital signs (4):** Maximum heart rate, minimum mean arterial pressure (MAP), maximum respiratory rate, and maximum temperature during the first 24 hours of ICU admission. Worst-case values were used consistent with severity scoring conventions (APACHE II, SOFA). MAP was selected over systolic and diastolic blood pressure due to its use as the primary hemodynamic target in the Surviving Sepsis Campaign (MAP \>65 mmHg) and its inclusion in SOFA scoring. SBP and DBP were excluded as multicollinear with MAP.

**Laboratory values (3):** Maximum creatinine, maximum white blood cell count (WBC), and minimum platelets during the first 24 hours. Bilirubin and lactate were excluded after sensitivity analysis showed no change in predictive performance (AUC 0.680 with and without), and both had \>15% missingness with patterns consistent with missing-not-at-random in ICU settings.

**Neurological (1):** Minimum Glasgow Coma Scale (GCS) total score during the first 24 hours, capturing neurological severity not represented by other predictors. This variable was computed as the sum of three subscores — eye opening (itemids 220739), verbal response (itemid 223900), and motor response (itemid 223901) — each recorded as the minimum value observed within the first 24 hours of ICU admission. The minimum subscores were used to capture the worst neurological state during the observation window, consistent with severity scoring conventions. The total GCS ranges from 3 (no response) to 15 (fully alert and oriented)."

**Demographics (3):** Age, gender (binary: male = 1), and race (collapsed to White, Black, Hispanic, Asian, Other, Unknown). \*Social determinants (1):\* Insurance type (Medicare, Medicaid, Private, Other, Unknown), which showed a 12% difference in prolonged stay prevalence across categories.

**Comorbidities (7)**: Hypertension, diabetes, chronic kidney disease (CKD), cardiovascular disease (CVD), chronic liver disease, immunosuppression, and vasopressor use — all derived from ICD-9 and ICD-10 secondary diagnosis codes.

**Interaction terms (3):** CKD×Diabetes (diabetic nephropathy), Chronic Liver×CVD (cardiohepatic syndrome), and GCS×Vasopressor (combined neurological and hemodynamic instability). These were included based on clinical reasoning that the joint presence of these conditions represents syndromes with distinct prognostic implications beyond their main effects.

**Pre-processing:** Outlier removal was performed by setting physiologically impossible values to missing (NA). Bounds were defined at biological limits — not clinical abnormality thresholds — to remove data entry errors while preserving true disease severity. Specific bounds applied were: heart rate (\<20 or \>300 bpm), mean arterial pressure (≤0 or \>200 mmHg), respiratory rate (\<5 or \>80 breaths/min), temperature (\<25°C or \>45°C), white blood cell count (\>500 K/µL), platelets (\>3,000 K/µL), creatinine (\>50 mg/dL), and Glasgow Coma Scale total (\<3 or \>15, as the scale is defined over this range by construction). Insurance type did not require outlier cleaning as it is categorical with fixed levels. Extreme values that were retained — such as respiratory rate of 73, WBC of 267, and temperature of 42.5°C — reflect physiologically possible values in severely ill sepsis patients. Following outlier removal, remaining missing values were imputed using the column median computed from the full cohort.

### Comorbidities

```{r}
#| label: comorbidities
cohort_ids <- sepsis_with_demo$hadm_id

# pull secondary diagnoses — both ICD-9 AND ICD-10

secondary_dx <- tbl(con_bq, "diagnoses_icd") |>
  filter(
    seq_num > 1,
    hadm_id %in% !!cohort_ids
  ) |>
  select(subject_id, hadm_id, icd_code, icd_version) |>
  collect()

comorbidity_flags <- secondary_dx |>
  mutate(
    # hypertension — ICD-10: I10 | ICD-9: 401-405
    hypertension = if_else(
      (icd_version == 10 & startsWith(icd_code, "I10")) |
      (icd_version == 9  & (startsWith(icd_code, "401") | startsWith(icd_code, "402") |
                             startsWith(icd_code, "403") | startsWith(icd_code, "404") |
                             startsWith(icd_code, "405"))),
      1L, 0L),

    # diabetes — ICD-10: E10/E11 | ICD-9: 250
    diabetes = if_else(
      (icd_version == 10 & (startsWith(icd_code, "E10") | startsWith(icd_code, "E11"))) |
      (icd_version == 9  &  startsWith(icd_code, "250")),
      1L, 0L),

    # chronic kidney disease — ICD-10: N18 | ICD-9: 585-586
    ckd = if_else(
      (icd_version == 10 & startsWith(icd_code, "N18")) |
      (icd_version == 9  & (startsWith(icd_code, "585") | startsWith(icd_code, "586"))),
      1L, 0L),

    # cardiovascular disease — ICD-10: I20/I21/I25/I48/I50/I70/I71
    #                           ICD-9: 410-414/427/428/440/441
    cvd = if_else(
      (icd_version == 10 & (
        startsWith(icd_code, "I20") | startsWith(icd_code, "I21") |
        startsWith(icd_code, "I25") | startsWith(icd_code, "I48") |
        startsWith(icd_code, "I50") | startsWith(icd_code, "I70") |
        startsWith(icd_code, "I71"))) |
      (icd_version == 9  & (
        startsWith(icd_code, "410") | startsWith(icd_code, "411") |
        startsWith(icd_code, "412") | startsWith(icd_code, "413") |
        startsWith(icd_code, "414") | startsWith(icd_code, "427") |
        startsWith(icd_code, "428") | startsWith(icd_code, "440") |
        startsWith(icd_code, "441"))),
      1L, 0L),

    # chronic liver disease — ICD-10: K70/K73/K74 | ICD-9: 571
    chronic_liver = if_else(
      (icd_version == 10 & (
        startsWith(icd_code, "K70") | startsWith(icd_code, "K73") |
        startsWith(icd_code, "K74"))) |
      (icd_version == 9  &  startsWith(icd_code, "571")),
      1L, 0L),

    # immunosuppression — ICD-10: D80/D81/D83/Z79/Z94/C81-C83/C91/C92
    #                     ICD-9: 279/V42/V45/200-202/204/205
    immunosuppression = if_else(
      (icd_version == 10 & (
        startsWith(icd_code, "D80") | startsWith(icd_code, "D81") |
        startsWith(icd_code, "D83") | startsWith(icd_code, "Z79") |
        startsWith(icd_code, "Z94") | startsWith(icd_code, "C81") |
        startsWith(icd_code, "C82") | startsWith(icd_code, "C83") |
        startsWith(icd_code, "C91") | startsWith(icd_code, "C92"))) |
      (icd_version == 9  & (
        startsWith(icd_code, "279") | startsWith(icd_code, "V42") |
        startsWith(icd_code, "V45") | startsWith(icd_code, "200") |
        startsWith(icd_code, "201") | startsWith(icd_code, "202") |
        startsWith(icd_code, "204") | startsWith(icd_code, "205"))),
      1L, 0L)
  ) |>
  group_by(subject_id, hadm_id) |>
  summarise(across(hypertension:immunosuppression, max), .groups = "drop") |>
  mutate(across(hypertension:immunosuppression, ~replace_na(.x, 0L)))

sepsis_with_comorbidities <- sepsis_with_demo |>
  left_join(comorbidity_flags, by = c("subject_id", "hadm_id")) |>
  mutate(across(hypertension:immunosuppression, ~replace_na(.x, 0L)))

# check 
cat("Cohort after comorbidity join: ", nrow(sepsis_with_comorbidities), "\n")
comorbidity_flags |>
  summarise(across(hypertension:immunosuppression, ~round(mean(.x)*100,1))) |>
  print(width=Inf)
```

### ICU Admission Window

```{r}
#| label: icu window
cohort_ids <- sepsis_with_comorbidities$hadm_id

# earliest ICU admission time per stay 
icu_admissions_bq <- tbl(con_bq, "icustays") |>
  filter(hadm_id %in% !!cohort_ids) |>
  group_by(subject_id, hadm_id) |>
  summarise(intime = min(intime, na.rm = TRUE), .groups = "drop")

icustays_raw <- icu_admissions_bq |> collect()
icu_admissions <- icustays_raw

cat("ICU stays pulled:", nrow(icustays_raw), "\n")
```

### Vital Signs

```{r}
#| label: vital-signs
# itemid 223761: convertss fahrenheit to celsius
vitals_raw <- tbl(con_bq, "chartevents") |>
  filter(
    itemid %in% c(220045, 220179, 220050, 220180, 220051,
                  220052, 220181, 220210, 223762, 223761),
    hadm_id %in% !!cohort_ids,
    !is.na(valuenum)
  ) |>
  select(subject_id, hadm_id, itemid, charttime, valuenum) |>
  inner_join(icu_admissions_bq, by = c("subject_id", "hadm_id")) |>
  filter(
    charttime >= intime,
    charttime <= sql("TIMESTAMP_ADD(intime, INTERVAL 24 HOUR)")
  ) |>
  mutate(
    valuenum = if_else(itemid == 223761, (valuenum - 32) * 5/9, valuenum)
  ) |>
  collect()   # collects only 24h window rows, not the whole table

# map itemids to vital names (some vitals have multiple itemids)
vital_items <- tibble::tibble(
  itemid     = c(220045, 220179, 220050, 220180, 220051,
                 220052, 220181, 220210, 223762, 223761),
  vital_name = c("heart_rate", "sbp",  "sbp",  "dbp",  "dbp",
                 "map",  "map",  "resp_rate", "temperature", "temperature")
)

# summarise each vital per admission (worst values)
vitals_summary <- vitals_raw |>
  left_join(vital_items, by = "itemid") |>
  group_by(subject_id, hadm_id, vital_name) |>
  summarise(
    value = case_when(
      # MAX — tachycardia indicates cardiac stress/infection severity
      first(vital_name) == "heart_rate"  ~ max(valuenum),
      
      # MIN — hypotension indicates shock/hemodynamic instability
      first(vital_name) == "sbp"         ~ min(valuenum),
      first(vital_name) == "dbp"         ~ min(valuenum),
      first(vital_name) == "map"         ~ min(valuenum),
      
      # MAX — tachypnea indicates respiratory distress/metabolic acidosis
      first(vital_name) == "resp_rate"   ~ max(valuenum),
      
      # MAX — peak fever indicates infection severity
      # (hypothermia in sepsis also dangerous but less common)
      first(vital_name) == "temperature" ~ max(valuenum),
    ),
    .groups = "drop"
  ) |>
  pivot_wider(names_from = vital_name, values_from = value)

cat("Patients with vitals:", nrow(vitals_summary), "\n")
```

### Lab Events

```{r}
#| label: laboratory-values
# creatinine (50912), WBC (51301), platelets (51265) only
# bilirubin and lactate excluded after sensitivity analysis showed
# no change in AUC (0.680 with and without)
labs_raw <- tbl(con_bq, "labevents") |>
  filter(
    itemid %in% c(50912, 51301, 51265),
    hadm_id %in% !!cohort_ids,
    !is.na(valuenum)
  ) |>
  select(subject_id, hadm_id, itemid, charttime, valuenum) |>
  inner_join(icu_admissions_bq, by = c("subject_id", "hadm_id")) |>
  filter(
    charttime >= intime,
    charttime <= sql("TIMESTAMP_ADD(intime, INTERVAL 24 HOUR)")
  ) |>
  collect()

lab_items <- tibble::tibble(
  itemid   = c(50912,        51301,  51265),
  lab_name = c("creatinine", "wbc",  "platelets")
)

labs_summary <- labs_raw |>
  left_join(lab_items, by = "itemid") |>
  group_by(subject_id, hadm_id, lab_name) |>
  summarise(
    value = case_when(
      first(lab_name) == "creatinine" ~ max(valuenum),
      first(lab_name) == "wbc"        ~ max(valuenum),
      first(lab_name) == "platelets"  ~ min(valuenum)
    ),
    .groups = "drop"
  ) |>
  pivot_wider(names_from = lab_name, values_from = value)

cat("Patients with labs:", nrow(labs_summary), "\n")
```

### Glasgow Coma Scale

```{r}
#| label: gcs
gcs_raw <- tbl(con_bq, "chartevents") |>
  filter(
    itemid %in% c(220739, 223900, 223901),
    hadm_id %in% !!cohort_ids,
    !is.na(valuenum)
  ) |>
  select(subject_id, hadm_id, itemid, charttime, valuenum) |>
  inner_join(icu_admissions_bq, by = c("subject_id", "hadm_id")) |>
  filter(
    charttime >= intime,
    charttime <= sql("TIMESTAMP_ADD(intime, INTERVAL 24 HOUR)")
  ) |>
  collect() |>
  group_by(subject_id, hadm_id, itemid) |>
  summarise(value = min(valuenum, na.rm = TRUE), .groups = "drop") |>
  pivot_wider(names_from = itemid, values_from = value) |>
  rename(gcs_eye    = `220739`,
         gcs_verbal = `223900`,
         gcs_motor  = `223901`) |>
  mutate(gcs_total = gcs_eye + gcs_verbal + gcs_motor) |>
  select(subject_id, hadm_id, gcs_total)

cat("Patients with GCS data:", nrow(gcs_raw), "\n")
```

### Vasopressor Use

```{r}
#| label: vasopressor use
# vasopressor itemids — norepinephrine, epinephrine, dopamine, vasopressin, phenylephrine
vasopressor_ids <- c(221906, 221289, 221662, 222315, 221749)

vasopressors_raw <- tbl(con_bq, "inputevents") |>
  filter(
    itemid %in% vasopressor_ids,
    hadm_id %in% !!cohort_ids
  ) |>
  select(subject_id, hadm_id, itemid, starttime) |>
  inner_join(icu_admissions_bq, by = c("subject_id", "hadm_id")) |>
  filter(
    starttime >= intime,
    starttime <= sql("TIMESTAMP_ADD(intime, INTERVAL 24 HOUR)")
  ) |>
  collect()

# binary flag — 1 if patient received vasopressors in first 24h, 0 otherwise
vasopressor_flags <- vasopressors_raw |>
  group_by(subject_id, hadm_id) |>
  summarise(vasopressor = 1L, .groups = "drop")

# check
cat("Patients with vasopressors:", nrow(vasopressor_flags), "\n")
```

### Final Dataset Assembly

```{r}
#| label: dataset-assembly
# safety step
sepsis_icu <- sepsis_with_comorbidities
# safety step
sepsis_with_vitals <- sepsis_icu |>
  left_join(vitals_summary, by = c("subject_id", "hadm_id"))

sepsis_with_labs <- sepsis_with_vitals |>
  left_join(labs_summary,      by = c("subject_id", "hadm_id")) |>
  left_join(vasopressor_flags, by = c("subject_id", "hadm_id")) |>
  left_join(gcs_raw,           by = c("subject_id", "hadm_id")) |>  # added GCS
  mutate(vasopressor = replace_na(vasopressor, 0L)) |>
  select(-sbp, -dbp)   # dropped sbp monitor 1, sdp monitor 2, dbp (both monitors), sbp multicollinear variables

cat("Final rows:", nrow(sepsis_with_labs), "\n")
cat("Vasopressor use (%):", round(mean(sepsis_with_labs$vasopressor)*100, 1), "\n")
cat("GCS missing (%):", round(mean(is.na(sepsis_with_labs$gcs_total))*100, 1), "\n")
```

### Outlier Removal

```{r}
#| label: outlier-removal
# goal: remove data entry errors, NOT filter out severely ill patients
sepsis_clean <- sepsis_with_labs |>
  mutate(
    # heart rate: no human survives >300, <20 is incompatible with life
    heart_rate  = if_else(heart_rate > 300 | heart_rate < 20,
                          NA_real_, heart_rate),

    # map: zero or negative impossible, >200 is data error
    map         = if_else(map <= 0 | map > 200, NA_real_, map),

    # resp rate: <5 impossible, >80 is data error
    resp_rate   = if_else(resp_rate < 5 | resp_rate > 80,
                          NA_real_, resp_rate),

    # temperature: <25 incompatible with life, >45 lethal/data error
    temperature = if_else(temperature < 25 | temperature > 45,
                          NA_real_, temperature),

    # wbc: >500 is beyond any recorded human value
    wbc         = if_else(wbc > 500, NA_real_, wbc),

    # platelets: >3000 is beyond any recorded human value
    platelets   = if_else(platelets > 3000, NA_real_, platelets),

    # creatinine: >50 is beyond any recorded human value
    creatinine  = if_else(creatinine > 50, NA_real_, creatinine),

    # GCS: scale is 3-15 by definition, outside = data error
    gcs_total   = if_else(gcs_total < 3 | gcs_total > 15,
                          NA_real_, gcs_total)
  )

# check missingness after cleaning
#sepsis_clean |>
#  summarise(across(everything(), ~ round(mean(is.na(.x)) * 100, 1))) |>
 # print(width = Inf)

# verify crazy values are gone
# summary(sepsis_clean[, c("resp_rate", "temperature",
                        #  "map", "wbc", "creatinine", "gcs_total")])

```

### Cohort Finalization and Imputation

```{r}
#| label: cohort finalization and imputation
# first admission per patient
sepsis_clean <- sepsis_clean |>
  arrange(subject_id, admittime) |>
  group_by(subject_id) |>
  slice(1) |>
  ungroup()

#cat("After first admission filter:", nrow(sepsis_clean), "\n")
#cat("Duplicate subject_ids:", sum(duplicated(sepsis_clean$subject_id)), "\n")

# remove early deaths (competing risk)
early_deaths <- tbl(con_bq, "admissions") |>
  select(subject_id, hadm_id, hospital_expire_flag, deathtime) |>
  collect()

sepsis_clean <- sepsis_clean |>
  left_join(early_deaths, by = c("subject_id", "hadm_id")) |>
  left_join(icustays_raw |> select(subject_id, hadm_id, intime),
            by = c("subject_id", "hadm_id")) |>
  filter(
    hospital_expire_flag == 0 |
    is.na(deathtime) |
    as.numeric(difftime(deathtime, intime, units = "hours")) > 48
  ) |>
  select(-hospital_expire_flag, -deathtime, -intime)

#cat("After removing early deaths:", nrow(sepsis_clean), "\n")

# median imputation
sepsis_imputed <- sepsis_clean |>
  mutate(across(
    .cols = c(heart_rate, map, resp_rate, temperature,
              creatinine, wbc, platelets, gcs_total),
    ~ if_else(is.infinite(.x), NA_real_, .x)
  )) |>
  mutate(across(
    .cols = c(heart_rate, map, resp_rate, temperature,
              creatinine, wbc, platelets, gcs_total),
    ~ coalesce(.x, median(.x, na.rm = TRUE))
  ))

cat("Final Analytic Cohort:", nrow(sepsis_imputed), "\n")
# cat("Duplicate subject_ids:", sum(duplicated(sepsis_imputed$subject_id)), "\n")

```

## Analytic Approach and Modeling

**Index time:** ICU admission. The prediction window was the first 24 hours of ICU admission, using worst-case values to capture peak physiological health status.

**Models:** Six classification models were trained and compared:

1.  *Logistic Regression* — baseline linear model
2.  *LASSO* (α = 1) — L1 penalization for automatic variable selection
3.  *Ridge* (α = 0) — L2 penalization retaining all predictors
4.  *Elastic Net* (α = 0.5) — combination of L1 and L2 penalties
5.  *Random Forest* — ensemble of decision trees with bootstrap aggregation
6.  *XGBoost* — gradient boosted trees with sequential error correction

All models were implemented in R using the `glmnet` package for regularized regression and the `tidymodels` framework for Random Forest (`ranger`) and XGBoost (`xgboost`).

### Model Matrix

```{r}
#| label: model-matrix
# race included as categorical predictor — dummy coded 
# gender recoded as binary (1 = M, 0 = F)

# define model variables FIRST
model_vars <- c("anchor_age", "gender", "race_collapsed",
                "heart_rate", "map", "resp_rate",
                "temperature", "creatinine",
                "wbc", "platelets", "gcs_total",
                "hypertension", "diabetes", "ckd", "cvd",
                "chronic_liver", "immunosuppression",
                "vasopressor", "insurance")

# THEN build model_df
model_df <- sepsis_imputed |>
  mutate(
    gender = if_else(gender == "M", 1L, 0L),
    race_collapsed = case_when(
      str_detect(race, "WHITE")    ~ "WHITE",
      str_detect(race, "BLACK")    ~ "BLACK",
      str_detect(race, "HISPANIC") ~ "HISPANIC",
      str_detect(race, "ASIAN")    ~ "ASIAN",
      race %in% c("UNKNOWN", "UNABLE TO OBTAIN",
                  "PATIENT DECLINED TO ANSWER") ~ "UNKNOWN",
      TRUE ~ "OTHER"
    ) |> as.factor(),
    insurance = if_else(is.na(insurance), "Unknown",
                        as.character(insurance)) |> as.factor()
  ) |>
  select(all_of(model_vars), prolonged_stay) |>
  mutate(across(where(is.numeric), ~if_else(is.infinite(.x), NA_real_, .x))) |>
  drop_na()

X <- model.matrix(~ . - prolonged_stay - 1, data = model_df)
y <- model_df$prolonged_stay

cat("Model matrix dimensions:", dim(X), "\n")
cat("Number of predictors:", ncol(X), "\n")

best_threshold <- function(roc_obj) {
  idx <- which.max(roc_obj$sensitivities + roc_obj$specificities - 1)
  roc_obj$thresholds[idx]
}

calc_metrics <- function(pred_prob, actual, threshold) {
  pred_class <- if_else(pred_prob >= threshold, 1L, 0L)
  sens <- mean(pred_class[actual == 1] == 1) * 100
  spec <- mean(pred_class[actual == 0] == 0) * 100
  c(sens, spec)
}

set.seed(42)
train_idx <- sample(1:nrow(X), 0.8 * nrow(X))
X_train <- X[train_idx, ]
X_test  <- X[-train_idx, ]
y_train <- y[train_idx]
y_test  <- y[-train_idx]

X_interact <- model.matrix(
  ~ . - prolonged_stay - 1 +
    chronic_liver:cvd +
    ckd:diabetes +
    gcs_total:vasopressor,
  data = model_df
)

X_train_i <- X_interact[train_idx, ]
X_test_i  <- X_interact[-train_idx, ]

cat("Training patients:", length(y_train), "\n")
cat("Test patients:    ", length(y_test), "\n")
```

### Logistic Regression, LASSO, Ridge, and Elastic Net

```{r}
#| label: regularized-models

# simple logistic regression (no regulization)
set.seed(42)
lr_model <- glm(y_train ~ ., data = as.data.frame(X_train_i), family = "binomial")
lr_pred  <- predict(lr_model, newdata = as.data.frame(X_test_i), type = "response")

# LASSO
set.seed(42)
cv_lasso    <- cv.glmnet(X_train_i, y_train, family = "binomial", alpha = 1, nfolds = 10)
lasso_model <- glmnet(X_train_i, y_train, family = "binomial", alpha = 1, lambda = cv_lasso$lambda.min)
lasso_pred  <- as.numeric(predict(lasso_model, newx = X_test_i, type = "response"))

# ridge
set.seed(42)
cv_ridge    <- cv.glmnet(X_train_i, y_train, family = "binomial", alpha = 0, nfolds = 10)
ridge_model <- glmnet(X_train_i, y_train, family = "binomial", alpha = 0, lambda = cv_ridge$lambda.min)
ridge_pred  <- as.numeric(predict(ridge_model, newx = X_test_i, type = "response"))

# elastic net
set.seed(42)
cv_enet    <- cv.glmnet(X_train_i, y_train, family = "binomial", alpha = 0.5, nfolds = 10)
enet_model <- glmnet(X_train_i, y_train, family = "binomial", alpha = 0.5, lambda = cv_enet$lambda.min)
enet_pred  <- as.numeric(predict(enet_model, newx = X_test_i, type = "response"))

# compute ROC objects
roc_lr    <- suppressMessages(roc(y_test, lr_pred))
roc_lasso <- suppressMessages(roc(y_test, lasso_pred))
roc_ridge <- suppressMessages(roc(y_test, ridge_pred))
roc_enet  <- suppressMessages(roc(y_test, enet_pred))

auc_lr    <- auc(roc_lr)
auc_lasso <- auc(roc_lasso)
auc_ridge <- auc(roc_ridge)
auc_enet  <- auc(roc_enet)

thresh_lr    <- best_threshold(roc_lr)
thresh_lasso <- best_threshold(roc_lasso)
thresh_ridge <- best_threshold(roc_ridge)
thresh_enet  <- best_threshold(roc_enet)

m_lr    <- calc_metrics(lr_pred,    y_test, thresh_lr)
m_lasso <- calc_metrics(lasso_pred, y_test, thresh_lasso)
m_ridge <- calc_metrics(ridge_pred, y_test, thresh_ridge)
m_enet  <- calc_metrics(enet_pred,  y_test, thresh_enet)

# LASSO selected predictors
lasso_coefs <- coef(lasso_model) |>
  as.matrix() |>
  as.data.frame() |>
  rownames_to_column("Variable") |>
  filter(Variable != "(Intercept)") |>
  mutate(dropped = if_else(s0 == 0, "DROPPED", "Kept"))

# Dropped Variables
cat("Dropped by LASSO:\n")
lasso_coefs |> filter(dropped == "DROPPED") |> pull(Variable) |> print()

cat("\nKept by LASSO:\n")
lasso_coefs |> filter(dropped == "Kept") |> 
  arrange(desc(abs(s0))) |> 
  pull(Variable) |> print()

lasso_coefs <- coef(lasso_model) |>
  as.matrix() |>
  as.data.frame() |>
  rownames_to_column("Variable") |>
  filter(Variable != "(Intercept)") |>
  print()
```

### Random Forest for Classification Model

```{r}
#| label: random-forest
# Random Forest 
set.seed(42)
sepsis_split <- initial_split(
  model_df |> mutate(prolonged_stay = as.factor(prolonged_stay)),
  prop = 0.8,
  strata = "prolonged_stay"
)
sepsis_other <- training(sepsis_split)
sepsis_test  <- testing(sepsis_split)

rf_recipe <- recipe(prolonged_stay ~ ., data = sepsis_other) |>
  step_dummy(all_nominal_predictors()) |>
  step_zv(all_numeric_predictors())

rf_mod <- rand_forest(
  mode  = "classification",
  mtry  = tune(),
  trees = tune()
) |>
  set_engine("ranger", importance = "impurity")

rf_wf <- workflow() |>
  add_recipe(rf_recipe) |>
  add_model(rf_mod)

# updated: trees expanded to 200-500 (plot showed upward trend at 300)
# mtry 1-5 unchanged — best mtry=3 was interior to range
param_grid_rf <- grid_regular(
  trees(range = c(200L, 500L)),
  mtry(range  = c(1L, 5L)),
  levels = c(4, 5)
)

set.seed(42)
folds <- vfold_cv(sepsis_other, v = 5, strata = "prolonged_stay")

rf_fit <- rf_wf |>
  tune_grid(
    resamples = folds,
    grid      = param_grid_rf,
    metrics   = metric_set(roc_auc, accuracy)
  )

# diagnosis plot
rf_fit |>
  collect_metrics() |>
  filter(.metric == "roc_auc") |>
  mutate(mtry = as.factor(mtry)) |>
  ggplot(aes(x = trees, y = mean, color = mtry, group = mtry)) +
  geom_line() + geom_point() +
  geom_errorbar(aes(ymin=mean-std_err, ymax=mean+std_err), width=10) +
  labs(x = "Number of Trees", y = "CV AUC",
       title = "Random Forest CV Tuning",
       subtitle = "Trees range expanded to 200-500 based on diagnostic plot") +
  theme_minimal()

rf_fit |> show_best(metric = "roc_auc")

best_rf      <- rf_fit |> select_best(metric = "roc_auc")
final_rf_wf  <- rf_wf  |> finalize_workflow(best_rf)
final_rf_fit <- final_rf_wf |> last_fit(sepsis_split)

final_rf_fit |> collect_metrics()

#final_rf_fit |>
#  extract_fit_parsnip() |>
 # vip(num_features = 19) +
 # labs(title = "Random Forest Variable Importance")

final_rf_fit |>
  collect_predictions() |>
  roc_curve(prolonged_stay, .pred_1) |>
  autoplot()

rf_auc <- final_rf_fit |>
  collect_metrics() |>
  filter(.metric == "roc_auc") |>
  pull(.estimate)

cat("Random Forest Test AUC:", round(rf_auc, 3), "\n")

```

### XG Boost Model

```{r}
#| label: xgboost
# XGBoost 
gb_recipe <- recipe(prolonged_stay ~ ., data = sepsis_other) |>
  step_dummy(all_nominal_predictors()) |>
  step_zv(all_numeric_predictors())

gb_mod <- boost_tree(
  mode       = "classification",
  trees      = 1000,
  tree_depth = tune(),
  learn_rate = tune()
) |>
  set_engine("xgboost")

gb_wf <- workflow() |>
  add_recipe(gb_recipe) |>
  add_model(gb_mod)

# updated: learn_rate narrowed from (-5, 2) to (-5, 0)
# diagnostic plot showed AUC crashes after learn_rate > 1
# concentrates all grid points in the effective region (1e-05 to 1)
param_grid_gb <- grid_regular(
  tree_depth(range = c(1L, 3L)),
  learn_rate(range = c(-5, 0), trans = log10_trans()),
  levels = c(3, 10)
)

set.seed(42)
folds <- vfold_cv(sepsis_other, v = 5, strata = "prolonged_stay")

gb_fit <- gb_wf |>
  tune_grid(
    resamples = folds,
    grid      = param_grid_gb,
    metrics   = metric_set(roc_auc, accuracy)
  )

# diagnosis plot
gb_fit |>
  collect_metrics() |>
  filter(.metric == "roc_auc") |>
  mutate(tree_depth = as.factor(tree_depth)) |>
  ggplot(aes(x = learn_rate, y = mean, color = tree_depth)) +
  geom_point() + geom_line() +
  scale_x_log10() +
  geom_errorbar(aes(ymin=mean-std_err, ymax=mean+std_err), width=0.1) +
  labs(x = "Learning Rate (log scale)", y = "CV AUC",
       title = "XGBoost CV Tuning",
       subtitle = "Learn rate narrowed to 1e-05 to 1 based on diagnostic plot") +
  theme_minimal()

gb_fit |> show_best(metric = "roc_auc")

best_gb      <- gb_fit |> select_best(metric = "roc_auc")
final_gb_wf  <- gb_wf  |> finalize_workflow(best_gb)
final_gb_fit <- final_gb_wf |> last_fit(sepsis_split)

final_gb_fit |> collect_metrics()

#final_gb_fit |>
#  extract_fit_parsnip() |>
#  vip(num_features = 19) +
#  labs(title = "XGBoost Variable Importance")

final_gb_fit |>
  collect_predictions() |>
  roc_curve(prolonged_stay, .pred_1) |>
  autoplot()

gb_auc <- final_gb_fit |>
  collect_metrics() |>
  filter(.metric == "roc_auc") |>
  pull(.estimate)

cat("XGBoost Test AUC:", round(gb_auc, 3), "\n")
```

## Model Evaluation

**Train/test split:** An 80/20 stratified split was used (n = 3,343 training, n = 836 test), stratified on the outcome to preserve class balance.

**Cross-validation:** 5-fold cross-validation was used for hyperparameter tuning of Random Forest (mtry, trees) and XGBoost (tree_depth, learn_rate).

**Hyperparameter tuning:** Initial tuning ranges were informed by diagnostic plots of CV AUC against hyperparameter values. For Random Forest, the trees range was expanded from 100–300 to 200–500 after diagnostic plots showed an upward trend at the boundary. For XGBoost, the learning rate range was narrowed from (1e-5, 100) to (1e-5, 1) after diagnostic plots revealed AUC collapsed for learn_rate \> 1, concentrating grid points in the effective region. Diagnostic plots confirmed that optimal hyperparameters were interior to the final search grids for both models.

```{r}
#| label: diagnostic-plots
# RF Diagnosis Plot — confirms trees range was appropriate
rf_fit |>
  collect_metrics() |>
  filter(.metric == "roc_auc") |>
  mutate(mtry = as.factor(mtry)) |>
  ggplot(aes(x=trees, y=mean, color=mtry, group=mtry)) +
  geom_line() + geom_point() +
  geom_errorbar(aes(ymin=mean-std_err, ymax=mean+std_err), width=0.2) +
  labs(title="Random Forest CV Tuning — Hyperparameter Diagnostic",
       subtitle="Best mtry=3 at trees=400, interior to grid — range was appropriate",
       x="Number of Trees", y="CV AUC") +
  theme_minimal()

# XGBoost Diagnosis Plot — confirmed learn_rate range needed narrowing
gb_fit |>
  collect_metrics() |>
  filter(.metric == "roc_auc") |>
  mutate(tree_depth = as.factor(tree_depth)) |>
  ggplot(aes(x=learn_rate, y=mean, color=tree_depth)) +
  geom_point() + geom_line() +
  scale_x_log10() +
  geom_errorbar(aes(ymin=mean-std_err, ymax=mean+std_err), width=0.1) +
  labs(title="XGBoost CV Tuning — Hyperparameter Diagnostic",
       subtitle="Peak at learn_rate~0.006, interior to narrowed range — grid was appropriate",
       x="Learning Rate (log scale)", y="CV AUC") +
  theme_minimal()
```

**Performance metrics:** Area under the ROC curve (AUC) was the primary metric. Decision thresholds were selected using the Youden index (maximizing sensitivity + specificity - 1). Sensitivity and specificity at the optimal threshold are reported for regularized regression models.

# Results

## Cohort Characteristics

The final analytic cohort comprised 4,179 adult sepsis ICU patients after applying all inclusion and exclusion criteria. Of these, 1,575 (37.7%) experienced prolonged ICU stays (≥14 days). The cohort had a median age of 65 years, was 57% male, and was predominantly White (61%). Medicare was the most common insurance type (52%). Hypertension (52%) and cardiovascular disease (38%) were the most prevalent comorbidities. Vasopressor use in the first 24 hours was recorded in 34% of patients.

```{r}
#| label: cohort-flow

# cohort flow - how many patients at each step
cohort_flow <- tibble(
  Step = c(
    "Adult sepsis ICU admissions",
    "After outcome filter (LOS > 0)",
    "After first admission filter",
    "After 48h early death exclusion",
    "Final analytic cohort"
  ),
  N = c(5017, 4953, 4478, 4179, 4179),
  label = c(
    "5,017",
    "4,953 (-64 same-day discharges)",
    "4,478 (-475 duplicate admissions)",
    "4,179 (-299 early deaths)",
    "4,179 (final)"
  )
) |>
  mutate(Step = factor(Step, levels = rev(Step)))

ggplot(cohort_flow, aes(x = N, y = Step)) +
  geom_col(fill = "steelblue", width = 0.5) +
  geom_text(aes(label = label), hjust = -0.05, size = 3.5) +
  labs(title = "Cohort Construction Flow",
       x = "Number of Patients", y = NULL) +
  xlim(0, 9000) +   # increased to fit long labels
  theme_minimal()

```

```{r}
#| label: outcome distribution
# sepsis_imputed: outcome distribution
sepsis_imputed |>
  count(prolonged_stay) |>
  mutate(label = if_else(prolonged_stay == 1, "Prolonged (>14 days)", "Normal (≤14 days)"),
         pct = round(n / sum(n) * 100, 1)) |>
  ggplot(aes(x = label, y = n, fill = label)) +
  geom_col(width = 0.5) +
  geom_text(aes(label = paste0(n, " (", pct, "%)")), 
            vjust = -0.5, size = 4) +
  scale_fill_manual(values = c("Prolonged (>14 days)" = "lightsalmon",
                               "Normal (≤14 days)"    = "lightblue")) +
  labs(title = "Outcome Distribution: Prolonged Hospital Stay",
       x = NULL, y = "Number of Patients") +
  theme_minimal() +
  theme(legend.position = "none")
```

```{r}
#| label: table1

# final cohort summary using gtsummary
sepsis_imputed |>
  mutate(
    prolonged_stay = factor(prolonged_stay, labels = c("Normal (<14 days)", "Prolonged (≥14 days)")),
    gender         = factor(gender, levels = c("F", "M"), 
                labels = c("Female", "Male")),
    insurance      = factor(insurance),
    race_collapsed = case_when(
      str_detect(race, "WHITE")    ~ "WHITE",
      str_detect(race, "BLACK")    ~ "BLACK",
      str_detect(race, "HISPANIC") ~ "HISPANIC",
      str_detect(race, "ASIAN")    ~ "ASIAN",
      race %in% c("UNKNOWN", "UNABLE TO OBTAIN",
                  "PATIENT DECLINED TO ANSWER") ~ "UNKNOWN",
      TRUE ~ "OTHER"
    ) |> as.factor()
  ) |>
  select(anchor_age, gender, race_collapsed, insurance,
         hypertension, diabetes, ckd, cvd, chronic_liver,
         immunosuppression, vasopressor, gcs_total,
         los_days, prolonged_stay) |>
  tbl_summary(
    by = prolonged_stay,
    statistic = list(
      all_continuous()  ~ "{median} ({p25}-{p75}); {mean} ({sd})",
      all_categorical() ~ "{n} ({p}%)"
    ),
    digits = list(all_continuous() ~ 1),
    label = list(
      anchor_age        ~ "Age (years)",
      gender            ~ "Gender",
      race_collapsed    ~ "Race",
      insurance         ~ "Insurance",
      hypertension      ~ "Hypertension",
      diabetes          ~ "Diabetes",
      ckd               ~ "Chronic Kidney Disease (CKD)",
      cvd               ~ "Cardiovascular Disease (CVD)",
      chronic_liver     ~ "Chronic Liver Disease",
      immunosuppression ~ "Immunosuppression",
      vasopressor       ~ "Vasopressor Use",
      gcs_total         ~ "Glasgow Coma Scale (GCS) Total",
      los_days          ~ "Length of Stay (LOS, days)"
    )
  ) |>
add_p() |>
  add_overall() |>
  bold_labels() |>
  modify_footnote(
    all_stat_cols() ~ "Continuous: Median (Q1-Q3); Mean (SD). Categorical: n (%)"
  )
```

```{r}
#| label: predictor-distributions
# Density Plot of continuous variables
sepsis_imputed |>
  pivot_longer(
    cols = c(gcs_total, creatinine, wbc,
             temperature, map, anchor_age,
             heart_rate, resp_rate, platelets),
    names_to  = "variable",
    values_to = "value"
  ) |>
  mutate(
    prolonged_stay = factor(prolonged_stay,
                            labels = c("Normal (<14 days)", "Prolonged (≥14 days)")),
    variable = recode(variable,
      gcs_total   = "Glasgow Coma Scale (GCS) Total",
      creatinine  = "Creatinine (mg/dL)",
      wbc         = "WBC (K/uL)",
      temperature = "Temperature (deg C)",
      map         = "MAP (mmHg)",
      anchor_age  = "Age (years)",
      heart_rate  = "Heart Rate (bpm)",
      resp_rate   = "Respiratory Rate (/min)",
      platelets   = "Platelets (K/uL)"
    )
  ) |>
  ggplot(aes(x = value, fill = prolonged_stay)) +
  geom_density(alpha = 0.5) +
  facet_wrap(~ variable, scales = "free", ncol = 3) +
  scale_fill_manual(values = c("aquamarine3", "salmon")) +
  labs(title = "Distribution of Key Continuous Predictors by Outcome",
       x = NULL, y = "Density", fill = "Outcome") +
  theme_minimal() +
  theme(legend.position = "bottom",
        strip.text = element_text(size = 9),
        panel.spacing = unit(1, "lines"))
```

```{r}
#| label: comorbidity-prevalence
# Comorbidity Prevalence by Outcome
sepsis_imputed |>
  mutate(prolonged_stay = factor(prolonged_stay,
                                  labels = c("Normal (<14 days)", 
                                             "Prolonged (≥14 days)"))) |>
  group_by(prolonged_stay) |>
  summarise(across(c(hypertension, diabetes, ckd, cvd,
                     chronic_liver, immunosuppression, vasopressor),
                   ~ round(mean(.x) * 100, 1))) |>
  pivot_longer(-prolonged_stay,
               names_to = "comorbidity",
               values_to = "pct") |>
  mutate(comorbidity = recode(comorbidity,
    hypertension      = "Hypertension",
    diabetes          = "Diabetes",
    ckd               = "Chronic Kidney Disease (CKD)",
    cvd               = "Cardiovascular Disease (CVD)",
    chronic_liver     = "Chronic Liver Disease",
    immunosuppression = "Immunosuppression",
    vasopressor       = "Vasopressor Use"
  )) |>
  ggplot(aes(x = reorder(comorbidity, pct), y = pct, fill = prolonged_stay)) +
  geom_col(position = "dodge", width = 0.6) +
  coord_flip() +
  scale_fill_manual(values = c("aquamarine3", "salmon")) +
  labs(title = "Comorbidity Prevalence by Outcome",
       x = NULL, y = "Prevalence (%)", fill = "Outcome") +
  theme_minimal() +
  theme(legend.position = "bottom")
```

## Model Performance

All six models demonstrated comparable discriminative performance on the held-out test set (AUC range 0.677–0.683). XGBoost achieved the highest test AUC (0.683) following diagnostic-guided hyperparameter refinement. Ridge regression (0.681) and LASSO (0.680) performed comparably. Random Forest achieved an AUC of 0.677, modestly below the regularized regression models. The narrow range of AUC values (0.006 difference between best and worst) suggests predominantly linear relationships between predictors and prolonged stay, and that increasing model complexity did not yield meaningful performance gains.

```{r}
#| label: model-performance-table
# full model comparison table
tibble(
  Model       = c("Logistic Regression", "LASSO", "Ridge", 
                  "Elastic Net", "Random Forest", "XGBoost"),
  AUC         = c(round(auc_lr,3), round(auc_lasso,3), 
                  round(auc_ridge,3), round(auc_enet,3),
                  round(rf_auc,3), round(gb_auc,3)),
  Threshold   = c(round(thresh_lr,3), round(thresh_lasso,3),
                  round(thresh_ridge,3), round(thresh_enet,3),
                  "—", "—"),
  Sensitivity = c(paste0(round(m_lr[1],1),"%"), 
                  paste0(round(m_lasso[1],1),"%"),
                  paste0(round(m_ridge[1],1),"%"), 
                  paste0(round(m_enet[1],1),"%"),
                  "—", "—"),
  Specificity = c(paste0(round(m_lr[2],1),"%"), 
                  paste0(round(m_lasso[2],1),"%"),
                  paste0(round(m_ridge[2],1),"%"), 
                  paste0(round(m_enet[2],1),"%"),
                  "—", "—")
) |>
  gt() |>
  tab_header(
    title    = "Model Performance on Held-Out Test Set",
    subtitle = "AUC, Youden-optimal threshold, sensitivity and specificity (n = 836 test patients)"
  ) |>
  tab_style(
    style     = list(cell_fill(color="#f0f4ff"),
                     cell_text(weight="bold")),
    locations = cells_body(rows = Model == "XGBoost")
  ) |>
  tab_style(
    style     = cell_fill(color="#f9f9f9"),
    locations = cells_body(rows = Model %in% 
                           c("Random Forest","XGBoost"))
  ) |>
  tab_footnote(
    footnote  = "Threshold and sensitivity/specificity not computed for tree-based models as they do not use a fixed linear threshold.",
    locations = cells_column_labels(columns = Threshold)
  ) |>
  cols_label(
    Model       = "Model",
    AUC         = "Test AUC",
    Threshold   = "Decision Threshold",
    Sensitivity = "Sensitivity",
    Specificity = "Specificity"
  )
```

```{r}
#| label: roc-curves
# build roc_df from test set ROC objects
lab_lr    <- paste0("Logistic Regression\nAUC = ", round(auc_lr, 3))
lab_ridge <- paste0("Ridge\nAUC = ", round(auc_ridge, 3))
lab_lasso <- paste0("LASSO\nAUC = ", round(auc_lasso, 3))
lab_enet  <- paste0("Elastic Net\nAUC = ", round(auc_enet, 3))

roc_df <- bind_rows(
  data.frame(FPR = 1 - roc_lr$specificities,    TPR = roc_lr$sensitivities,    Model = lab_lr),
  data.frame(FPR = 1 - roc_ridge$specificities,  TPR = roc_ridge$sensitivities,  Model = lab_ridge),
  data.frame(FPR = 1 - roc_lasso$specificities,  TPR = roc_lasso$sensitivities,  Model = lab_lasso),
  data.frame(FPR = 1 - roc_enet$specificities,   TPR = roc_enet$sensitivities,   Model = lab_enet)
)

# combined all 4 into one graph
ggplot(roc_df, aes(x = FPR, y = TPR, color = Model)) +
  geom_line(lwd = 1) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "gray50") +
  scale_color_manual(values = setNames(
    c("yellow", "blue", "green", "red"),
    c(lab_lr, lab_ridge, lab_lasso, lab_enet)
  )) +
  labs(
    title = "ROC Curves by Model",
    x     = "False Positive Rate (1 - Specificity)",
    y     = "True Positive Rate (Sensitivity)"
  ) +
  theme_minimal() +
  theme(legend.position = "bottom")
```

The narrow range of AUC values across all models (0.006 difference between best and worst) suggests that the relationships between predictors and prolonged stay are predominantly linear, and that increasing model complexity did not yield meaningful performance gains in this dataset.

## Feature Importance

LASSO variable selection retained 24 of 30 model matrix columns, dropping race_collapsedASIAN, race_collapsedUNKNOWN, race_collapsedWHITE, diabetes (main effect), immunosuppression, and insurancePrivate. Notably, all three interaction terms were retained, with the CKD×Diabetes interaction kept while diabetes as a main effect was dropped — consistent with diabetic nephropathy as the clinically meaningful entity. Chronic liver disease and cardiovascular disease had the largest positive LASSO coefficients. Insurance: Unknown had the largest negative coefficient, likely reflecting documentation patterns rather than true clinical differences.

```{r}
#| label: table2
# LASSO coefficient table for report
coef(lasso_model) |>
  as.matrix() |>
  as.data.frame() |>
  rownames_to_column("Variable") |>
  filter(Variable != "(Intercept)", s0 != 0) |>
  mutate(
    Variable = recode(Variable,
      "anchor_age"             = "Age",
      "gender"                 = "Gender (Male vs Female)",
      "heart_rate"             = "Heart Rate (HR)",
      "map"                    = "Mean Arterial Pressure (MAP)",
      "resp_rate"              = "Respiratory Rate (RR)",
      "temperature"            = "Temperature",
      "creatinine"             = "Creatinine",
      "wbc"                    = "White Blood Cell Count (WBC)",
      "platelets"              = "Platelets",
      "gcs_total"              = "Glasgow Coma Scale (GCS)",
      "hypertension"           = "Hypertension",
      "diabetes"               = "Diabetes",
      "ckd"                    = "Chronic Kidney Disease (CKD)",
      "cvd"                    = "Cardiovascular Disease (CVD)",
      "chronic_liver"          = "Chronic Liver Disease",
      "immunosuppression"      = "Immunosuppression",
      "vasopressor"            = "Vasopressor Use",
      "race_collapsedBLACK"    = "Race: Black (vs Asian)",
      "race_collapsedHISPANIC" = "Race: Hispanic (vs Asian)",
      "race_collapsedOTHER"    = "Race: Other (vs Asian)",
      "race_collapsedUNKNOWN"  = "Race: Unknown (vs Asian)",
      "race_collapsedWHITE"    = "Race: White (vs Asian)",
      "insuranceMedicare"      = "Insurance: Medicare (vs Medicaid)",
      "insuranceOther"         = "Insurance: Other (vs Medicaid)",
      "insurancePrivate"       = "Insurance: Private (vs Medicaid)",
      "insuranceUnknown"       = "Insurance: Unknown (vs Medicaid)",
      "chronic_liver:cvd"      = "Interaction: Chronic Liver x CVD",
      "cvd:chronic_liver"      = "Interaction: Chronic Liver x CVD",
      "ckd:diabetes"           = "Interaction: CKD x Diabetes",
      "diabetes:ckd"           = "Interaction: CKD x Diabetes",
      "gcs_total:vasopressor"  = "Interaction: GCS x Vasopressor"
    ),
    Direction = if_else(s0 > 0, "Increases Risk", "Decreases Risk"),
    s0        = round(s0, 4)
  ) |>
  arrange(desc(s0)) |>
  rename(Coefficient = s0) |>
  gt() |>
  tab_header(
    title    = "LASSO Logistic Regression Coefficients",
    subtitle = "Predictors of prolonged ICU stay (>=14 days) — non-zero coefficients only"
  ) |>
  tab_style(
    style     = cell_fill(color="#fff0f0"),
    locations = cells_body(rows = Direction == "Increases Risk")
  ) |>
  tab_style(
    style     = cell_fill(color="#f0f8ff"),
    locations = cells_body(rows = Direction == "Decreases Risk")
  ) |>
  tab_style(
    style     = cell_text(weight="bold"),
    locations = cells_body(rows = abs(Coefficient) > 0.2)
  ) |>
  cols_label(
    Variable    = "Predictor",
    Coefficient = "LASSO Coefficient",
    Direction   = "Direction"
  ) |>
  tab_footnote(
    footnote  = "Reference categories: Race = Asian; Insurance = Medicaid. Variables with zero coefficient were dropped by LASSO penalization.",
    locations = cells_column_labels(columns = Variable)
  )
```

### VIP Alignment

Random Forest and XGBoost variable importance plots both identified GCS total as the top predictor by importance magnitude, followed by age and platelets. This is consistent with LASSO findings where GCS had the largest coefficient among continuous clinical predictors. The alignment across all three modeling approaches strengthens confidence in GCS as a robust predictor of prolonged ICU stay.

```{r}
#| label: vip-plots

# Random Forest Variable Importance
final_rf_fit |>
  extract_fit_parsnip() |>
  vip(num_features = 19) +
  labs(
    title    = "Random Forest Variable Importance",
    subtitle = "Top 19 predictors ranked by mean decrease in impurity",
    x        = "Importance",
    y        = NULL
  ) +
  theme_minimal()

# XGBoost Variable Importance  
final_gb_fit |>
  extract_fit_parsnip() |>
  vip(num_features = 19) +
  labs(
    title    = "XGBoost Variable Importance",
    subtitle = "Top 19 predictors ranked by gain",
    x        = "Importance",
    y        = NULL
  ) +
  theme_minimal()
```

# Discussion

## Summary of Findings

This study developed and evaluated six machine learning models for predicting prolonged ICU stay in adult sepsis patients using MIMIC-IV. All models performed comparably (AUC 0.677–0.683), with XGBoost achieving the highest test AUC (0.683). The convergence of performance across model classes — from simple logistic regression to gradient boosted trees — suggests that the predictive signal in this dataset is largely captured by linear relationships, and that additional model complexity does not provide meaningful clinical benefit. Comorbidity burden, particularly chronic liver disease and cardiovascular disease, emerged as the strongest predictors of prolonged stay, more so than acute physiological derangements. This finding suggests that pre-admission health status is a key driver of extended ICU stay in sepsis patients, and that risk stratification should account for chronic disease burden alongside acute severity.

## Comparison with Previous Work

Prior studies applying machine learning to ICU outcome prediction have similarly found that regularized regression methods perform comparably to more complex ensemble approaches on structured tabular clinical data. The predominance of comorbidity burden over acute physiology in predicting prolonged stay is consistent with the broader literature on post-acute care outcomes in critically ill patients.

## Strengths and Limitations

\*Strengths:\*\* This study used a large, well-validated clinical database with comprehensive coverage of diagnoses, vital signs, laboratory values, and administrative data. Both ICD-9 and ICD-10 codes were used for comorbidity identification, ensuring coverage across the full temporal range of the dataset. Clinically motivated interaction terms were pre-specified based on known clinical syndromes, and their retention by LASSO provides data-driven validation of their relevance.

**Limitations:** The study was conducted at a single center (BIDMC) and results may not generalize to other institutions or patient populations. The outcome was defined using hospital LOS as a proxy for ICU LOS, which may introduce misclassification for patients transferred between units. Median imputation was used for missing values, which may underestimate uncertainty. The model uses first-admission worst-case values and does not capture dynamic changes in patient status over time. AUC of 0.680 reflects moderate discriminative ability — the model should supplement, not replace, clinical judgment.

## Implications

These findings have several practical implications. First, the strong role of comorbidity burden suggests that sepsis patients with chronic liver disease or cardiovascular disease should be flagged early for proactive discharge planning and care coordination. Second, the comparable performance of LASSO and XGBoost supports the use of interpretable models in clinical decision support applications, where transparency is critical for clinician trust and adoption. Third, the inclusion of insurance type and race as independent predictors highlights the role of social determinants of health in ICU outcomes, suggesting that equitable care initiatives should consider these factors in resource allocation.

# Conclusion

This study demonstrates that prolonged ICU stay in sepsis patients can be predicted with moderate accuracy (AUC 0.683) using routinely collected clinical and administrative data from the first 24 hours of ICU admission. All six machine learning models performed comparably, suggesting predominantly linear relationships between predictors and outcome. Comorbidity burden — particularly chronic liver disease and cardiovascular disease — was the strongest driver of prolonged stay, more so than acute physiological derangements. These findings support the use of interpretable models such as LASSO for clinical deployment, and highlight the importance of pre-admission health status in early risk stratification for sepsis patients. Future work could extend this approach by incorporating dynamic time-series data from the full ICU stay, validating the model in external cohorts, and exploring fairness-aware modeling approaches to address the observed associations between race, insurance, and outcomes.

# References

# AI Usage Acknowledgement

Claude Sonnet 4.6 (Anthropic) was used throughout the project to assist with R code debugging, Shiny application development, and grammatical editing of report text. All analytical decisions, clinical reasoning, model specifications, and final written content were reviewed, verified, and directed by the authors. The MIMIC-IV data pipeline, data engineering, model training, and the results reported reflect the authors' own analytical process.

```{r}
#| label: save-app-data

#saving to rds file
app_data <- list(
  cohort      = sepsis_imputed,
  lasso_model = lasso_model
)

saveRDS(app_data, "/Users/charlottelee/Documents/biostat-203b-2026-winter/final_project/app_data.rds")
cat("Saved! File size:", 
    round(file.size("/Users/charlottelee/Documents/biostat-203b-2026-winter/final_project/app_data.rds")/1e6, 1), 
    "MB\n")
```
