---
title: "Biostat Final Project"
subtitle: "Due Mar 18 @ 11:59PM"
author: "Charlotte Lee and 206782165"
execute:
  cache: true         
format:
  html:
    theme: cosmo
    embed-resources: true
    number-sections: false
    toc: true
    toc-depth: 4
    toc-location: left
    code-fold: false
---

## Load Libraries

```{r}
#| message: false
#| warning: false
library(bigrquery)
library(dbplyr)
library(DBI)
library(gt)
library(gtsummary)
library(tidyverse)
library(glmnet)
library(pROC)
library(corrplot)
```

## Connect to The Data

```{r}
# authenticate
bq_auth(path = "biostat-203b-2026-winter-92fefbfab477.json")

# connect
con_bq <- dbConnect(
    bigrquery::bigquery(),
    project = "biostat-203b-2025-winter",
    dataset = "mimiciv_3_1",
    billing = "biostat-203b-2025-winter"
)
con_bq
```

# Data Cleaning + Pre-Processing

## 1. Filtering out Age (\>=18) & Primary Sepsis Diagnosis (upon admission)

```{r}
# load raw tables
diagnoses_icd_tbl <- tbl(con_bq, "diagnoses_icd") |>
  arrange(subject_id, hadm_id)

d_icd_diagnoses_tbl <- tbl(con_bq, "d_icd_diagnoses")

patients_tbl <- tbl(con_bq, "patients") |>
  select(subject_id, anchor_age, gender)

# match sepsis icd codes from lookup
sepsis_codes_tbl <- d_icd_diagnoses_tbl |>
  filter(sql("LOWER(long_title) LIKE '%sepsis%'")) |>
  select(icd_code, icd_version, long_title)

sepsis_codes_tbl |> collect() |> print(n = 50)

# keep only primary diagnosis (seq_num == 1)
sepsis_cohort_tbl <- diagnoses_icd_tbl |>
  filter(seq_num == 1) |>
  inner_join(sepsis_codes_tbl, by = c("icd_code", "icd_version")) |>
  select(subject_id, hadm_id, seq_num, icd_code, icd_version, long_title)

# filter adults only
adults_tbl <- patients_tbl |>
  filter(anchor_age >= 18)

sepsis_adult_cohort_tbl <- sepsis_cohort_tbl |>
  inner_join(adults_tbl, by = "subject_id")

cohort_df <- sepsis_adult_cohort_tbl |> collect()

cat("Total admissions: ", nrow(cohort_df), "\n")
cat("Unique patients:  ", n_distinct(cohort_df$subject_id), "\n")
print(summary(cohort_df$anchor_age))

cohort_df |>
  count(icd_code, icd_version, long_title, sort = TRUE) |>
  print(n = 5)


# check:
cat("Total admissions: ", nrow(cohort_df), "\n")
cat("Unique patients:  ", n_distinct(cohort_df$subject_id), "\n")

cat("\n--- Age Summary ---\n")
print(summary(cohort_df$anchor_age))

```

## 2. Outcome Variable

Outcome (**admissions.csv**)

```{r}
# pulls LOS + race in one query instead of two
admissions_tbl <- tbl(con_bq, "admissions") |>
  select(subject_id, hadm_id, admittime, dischtime, race) |>
  mutate(los_days = sql("TIMESTAMP_DIFF(dischtime, admittime, DAY)")) |>
  collect()

# join onto cohort and create outcome variable (prolonged stay >= 14 days)
sepsis_with_outcome <- cohort_df |>
  inner_join(admissions_tbl, by = c("subject_id", "hadm_id")) |>
  filter(!is.na(los_days), los_days > 0) |>
  mutate(prolonged_stay = if_else(los_days >= 14, 1L, 0L))

# check:
cat("Total admissions:       ", nrow(sepsis_with_outcome), "\n")
cat("Unique patients:        ", n_distinct(sepsis_with_outcome$subject_id), "\n")
cat("Prolonged stay (>=14d): ", sum(sepsis_with_outcome$prolonged_stay), "\n")
cat("% prolonged:            ", round(mean(sepsis_with_outcome$prolonged_stay)*100, 1), "%\n")
cat("Median LOS:             ", median(sepsis_with_outcome$los_days), "days\n")

```

## 3. Adding Predictors

### 3.1 Demographics

```{r}
# race already joined in section 2 — no second BQ pull needed
sepsis_with_demo <- sepsis_with_outcome

# check:
cat("Rows after joining demographics: ", nrow(sepsis_with_demo), "\n")
glimpse(sepsis_with_demo)
```

### 3.2 Comorbidities

```{r}
cohort_ids <- sepsis_with_demo$hadm_id

# pull secondary diagnoses and flag comorbidities using ICD-10 codes
secondary_dx <- tbl(con_bq, "diagnoses_icd") |>
  filter(
    seq_num > 1,
    icd_version == 10,
    hadm_id %in% !!cohort_ids
  ) |>
  select(subject_id, hadm_id, icd_code) |>
  collect()

comorbidity_flags <- secondary_dx |>
  mutate(
    hypertension      = if_else(startsWith(icd_code, "I10"), 1L, 0L),
    diabetes          = if_else(
                          startsWith(icd_code, "E10") |
                          startsWith(icd_code, "E11"), 1L, 0L),
    ckd               = if_else(startsWith(icd_code, "N18"), 1L, 0L),
    cvd               = if_else(
                          startsWith(icd_code, "I20") |
                          startsWith(icd_code, "I21") |
                          startsWith(icd_code, "I25") |
                          startsWith(icd_code, "I48") |
                          startsWith(icd_code, "I50") |
                          startsWith(icd_code, "I70") |
                          startsWith(icd_code, "I71"), 1L, 0L),
    chronic_liver     = if_else(
                          startsWith(icd_code, "K70") |
                          startsWith(icd_code, "K73") |
                          startsWith(icd_code, "K74"), 1L, 0L),
    immunosuppression = if_else(
                          startsWith(icd_code, "D80") |
                          startsWith(icd_code, "D81") |
                          startsWith(icd_code, "D83") |
                          startsWith(icd_code, "Z79") |
                          startsWith(icd_code, "Z94") |
                          startsWith(icd_code, "C81") |
                          startsWith(icd_code, "C82") |
                          startsWith(icd_code, "C83") |
                          startsWith(icd_code, "C91") |
                          startsWith(icd_code, "C92"), 1L, 0L)
  ) |>
  group_by(subject_id, hadm_id) |>
  summarise(across(hypertension:immunosuppression, max), .groups = "drop") |>
  mutate(across(hypertension:immunosuppression, ~replace_na(.x, 0L)))

sepsis_with_comorbidities <- sepsis_with_demo |>
  left_join(comorbidity_flags, by = c("subject_id", "hadm_id")) |>
  mutate(across(hypertension:immunosuppression, ~replace_na(.x, 0L)))

# check:
cat("Rows after joining comorbidities: ", nrow(sepsis_with_comorbidities), "\n")
comorbidity_flags |>
  summarise(across(hypertension:immunosuppression, ~round(mean(.x)*100,1))) |>
  print()
dim(sepsis_with_comorbidities)

```

### 3.3 Comorbidities

```{r}
# safety step: ensure we pull right admissions
cohort_ids <- sepsis_with_comorbidities$hadm_id

# pull earliest ICU admission time per stay
icu_admissions_bq <- tbl(con_bq, "icustays") |>
  filter(hadm_id %in% !!cohort_ids) |>
  group_by(subject_id, hadm_id) |>
  summarise(intime = min(intime, na.rm = TRUE), .groups = "drop")

# keep a local copy too for later filtering
icustays_raw <- icu_admissions_bq |> collect()
icu_admissions <- icustays_raw  # alias used downstream

#check
cat("ICU stays pulled:", nrow(icustays_raw), "\n")

```

### 3.4 Vital Items

```{r}
# pull vitals from chartevents within first 24h of ICU admission
# itemid 223761 is fahrenheit so converting to celsius
vitals_raw <- tbl(con_bq, "chartevents") |>
  filter(
    itemid %in% c(220045, 220179, 220050, 220180, 220051,
                  220052, 220181, 220210, 223762, 223761),
    hadm_id %in% !!cohort_ids,
    !is.na(valuenum)
  ) |>
  select(subject_id, hadm_id, itemid, charttime, valuenum) |>
  inner_join(icu_admissions_bq, by = c("subject_id", "hadm_id")) |>
  filter(
    charttime >= intime,
    charttime <= sql("TIMESTAMP_ADD(intime, INTERVAL 24 HOUR)")
  ) |>
  mutate(
    valuenum = if_else(itemid == 223761, (valuenum - 32) * 5/9, valuenum)
  ) |>
  collect()   # now collects only 24h window rows, not the whole table

# map itemids to vital names (some vitals have multiple itemids)
vital_items <- tibble::tibble(
  itemid     = c(220045, 220179, 220050, 220180, 220051,
                 220052, 220181, 220210, 223762, 223761),
  vital_name = c("heart_rate", "sbp",  "sbp",  "dbp",  "dbp",
                 "map",  "map",  "resp_rate", "temperature", "temperature")
)

# summarise each vital per admission (worst values)
vitals_summary <- vitals_raw |>
  left_join(vital_items, by = "itemid") |>
  group_by(subject_id, hadm_id, vital_name) |>
  summarise(
    value = case_when(
      first(vital_name) == "heart_rate"  ~ max(valuenum),
      first(vital_name) == "sbp"         ~ min(valuenum),
      first(vital_name) == "dbp"         ~ min(valuenum),
      first(vital_name) == "map"         ~ min(valuenum),
      first(vital_name) == "resp_rate"   ~ max(valuenum),
      first(vital_name) == "temperature" ~ max(valuenum),
    ),
    .groups = "drop"
  ) |>
  pivot_wider(names_from = vital_name, values_from = value)

# check
cat("Rows after joining vitals:", nrow(vitals_summary), "\n")

# may take up to a minute to run
```

### 3.5 Lab Events

```{r}
#Confirm Lab Items: 
# verify all lab itemids are correct
#tbl(con_bq, "d_labitems") |>
 # filter(itemid %in% c(50912, 50885, 51301, 51265, 50813)) |>
#  select(itemid, label) |>
#  arrange(itemid) |>
#  collect() |>
#  print()

# pull labs from labevents within first 24h of ICU admission
labs_raw <- tbl(con_bq, "labevents") |>
  filter(
    itemid %in% c(50912, 50885, 51301, 51265, 50813),
    hadm_id %in% !!cohort_ids,
    !is.na(valuenum)
  ) |>
  select(subject_id, hadm_id, itemid, charttime, valuenum) |>
  collect() |>
  inner_join(icu_admissions, by = c("subject_id", "hadm_id")) |>
  filter(
    charttime >= intime,
    charttime <= intime + lubridate::hours(24)
  )

# map itemids to lab names
lab_items <- tibble::tibble(
  itemid   = c(50912,        50885,       51301,  51265,       50813),
  lab_name = c("creatinine", "bilirubin", "wbc",  "platelets", "lactate")
)

# worst values per admission
labs_summary <- labs_raw |>
  left_join(lab_items, by = "itemid") |>
  group_by(subject_id, hadm_id, lab_name) |>
  summarise(
    value = case_when(
      first(lab_name) == "creatinine" ~ max(valuenum),
      first(lab_name) == "bilirubin"  ~ max(valuenum),
      first(lab_name) == "wbc"        ~ max(valuenum),
      first(lab_name) == "platelets"  ~ min(valuenum),
      first(lab_name) == "lactate"    ~ max(valuenum),
    ),
    .groups = "drop"
  ) |>
  pivot_wider(names_from = lab_name, values_from = value)

# check
cat("Rows after joining labs:", nrow(labs_summary), "\n")
```

### 3.6 Subset to CU Patients only

```{r}
# restrict to ICU patients only then join vitals and labs
sepsis_icu <- sepsis_with_comorbidities |>
  filter(hadm_id %in% icustays_raw$hadm_id)

sepsis_with_vitals <- sepsis_icu |>
  left_join(vitals_summary, by = c("subject_id", "hadm_id"))

sepsis_with_labs <- sepsis_with_vitals |>
  left_join(labs_summary, by = c("subject_id", "hadm_id"))

```

## 4. Raw Data

```{r}
#raw data view
summary(sepsis_with_labs)

# check
cat("raw final dataset:", nrow(sepsis_icu), "\n")
cat("Final rows:", nrow(sepsis_with_labs), "\n")

# check % missingness across all columns
sepsis_with_labs %>%
  summarise(across(everything(), ~ round(mean(is.na(.x)) * 100, 1))) %>%
  print(width = Inf)


```

## 5. Outlier Cleaning

```{r}
# remove physiologically impossible values for vitals and labs 
sepsis_clean <- sepsis_with_labs |>
  mutate(
    heart_rate  = if_else(heart_rate > 300, NA_real_, heart_rate),
    sbp         = if_else(sbp < 0 | sbp > 300, NA_real_, sbp),
    dbp         = if_else(dbp < 0 | dbp > 200, NA_real_, dbp),
    map         = if_else(map < 0 | map > 200, NA_real_, map),
    resp_rate   = if_else(resp_rate > 60, NA_real_, resp_rate),
    temperature = if_else(temperature < 25 | temperature > 45, NA_real_, temperature),
    lactate     = if_else(lactate > 30, NA_real_, lactate),
    wbc         = if_else(wbc > 200, NA_real_, wbc)
  )

# check % missingness after cleaning
  # missingness increases slightly after cleaning — expected, since impossible values
  # are now correctly coded as NA rather than real data
      #ex: map goes from 0.4->1.4%     
      #meaning 0.7% of map values were outside 0-200 range & got replaced with NA
sepsis_clean |>
  summarise(across(everything(), ~ round(mean(is.na(.x)) * 100, 1))) %>%
  print(width = Inf)
```

## 6. Imputation

```{r}
# aggregate per patient then impute remaining NAs with median
sepsis_imputed <- sepsis_clean %>%
  group_by(subject_id) %>%
  summarise(
    # vitals: mean per patient to reduce extreme spikes
    heart_rate  = mean(heart_rate, na.rm = TRUE),
    sbp         = mean(sbp, na.rm = TRUE),
    dbp         = mean(dbp, na.rm = TRUE),
    map         = mean(map, na.rm = TRUE),
    resp_rate   = mean(resp_rate, na.rm = TRUE),
    temperature = mean(temperature, na.rm = TRUE),
    
    # labs: max per patient to capture peak severity (min for platelets — low is worse)
    creatinine  = max(creatinine, na.rm = TRUE),
    bilirubin   = max(bilirubin, na.rm = TRUE),
    wbc         = max(wbc, na.rm = TRUE),
    lactate     = max(lactate, na.rm = TRUE),
    
    #needs to be min for platelets
    platelets   = min(platelets, na.rm = TRUE),
    
    # first() keeps the columns/variables 
    # colesce() fills with median imputation
    # demographics, comorbidities, outcome
    anchor_age        = first(anchor_age),
    gender            = first(gender),
    race              = first(race),
    hypertension      = first(hypertension),
    diabetes          = first(diabetes),
    ckd               = first(ckd),
    cvd               = first(cvd),
    chronic_liver     = first(chronic_liver),
    immunosuppression = first(immunosuppression),
    prolonged_stay    = first(prolonged_stay),
    los_days          = first(los_days)
  ) %>%
  ungroup() %>%
  mutate(across(
    .cols = c(heart_rate, sbp, dbp, map, resp_rate, temperature,
              creatinine, bilirubin, wbc, platelets, lactate),
    ~ coalesce(.x, median(.x, na.rm = TRUE))
  ))

# confirm no remaining missingness
sepsis_imputed %>%
  summarise(across(c(heart_rate, sbp, dbp, map, resp_rate, temperature,
                     creatinine, bilirubin, wbc, platelets, lactate),
                   ~ round(mean(is.na(.x)) * 100, 1))) %>%
  print()


```

## 7. Final Cohort

```{r}
# final cohort summary
cat("Total patients:", nrow(sepsis_imputed), "\n")
cat("Unique subjects:", n_distinct(sepsis_imputed$subject_id), "\n")

# anchor age predictor: summary statistics
summary(sepsis_imputed$anchor_age)

# count of gender
table(sepsis_imputed$gender)

# count of prolonged stay
table(sepsis_imputed$prolonged_stay)

# proportion of prolonged stay
prop.table(table(sepsis_imputed$prolonged_stay)) |> round(3)

# outcome: summary statistics
summary(sepsis_imputed$los_days)
```

## 8. Graphs & Tables

```{r}
# raw data: sepsis_with_labs --> outcome distribution
sepsis_with_labs |>
    count(prolonged_stay) |>
    mutate(label = if_else(prolonged_stay == 1, "Prolonged (>14 days)", "Normal (≤14 days)"),
           pct = round(n / sum(n) * 100, 1)) |>
    ggplot(aes(x = label, y = n, fill = label)) +
    geom_col(width = 0.5) +
    geom_text(aes(label = paste0(n, " (", pct, "%)")), 
              vjust = -0.5, size = 4) +
    scale_fill_manual(values = c("Prolonged (>14 days)" = "salmon",
                                 "Normal (≤14 days)"    = "aquamarine2")) +
    labs(title = "Outcome Distribution: Prolonged Hospital Stay",
         x = NULL, y = "Number of Patients") +
    theme_minimal() +
    theme(legend.position = "none")

# sepsis_imputed: outcome distribution
sepsis_imputed |>
    count(prolonged_stay) |>
    mutate(label = if_else(prolonged_stay == 1, "Prolonged (>14 days)", "Normal (≤14 days)"),
           pct = round(n / sum(n) * 100, 1)) |>
    ggplot(aes(x = label, y = n, fill = label)) +
    geom_col(width = 0.5) +
    geom_text(aes(label = paste0(n, " (", pct, "%)")), 
              vjust = -0.5, size = 4) +
    scale_fill_manual(values = c("Prolonged (>14 days)" = "salmon",
                                 "Normal (≤14 days)"    = "aquamarine")) +
    labs(title = "Outcome Distribution: Prolonged Hospital Stay",
         x = NULL, y = "Number of Patients") +
    theme_minimal() +
    theme(legend.position = "none")

# sepsis_imputed: age distribution
ggplot(sepsis_imputed, aes(x = anchor_age)) +
    geom_histogram(binwidth = 5, fill = "lightblue", color = "white") +
    geom_vline(xintercept = mean(sepsis_imputed$anchor_age), 
               linetype = "dashed", color = "red", linewidth = 0.8) +
    annotate("text", x = mean(sepsis_imputed$anchor_age) + 2, y = 300,
             label = paste0("Mean: ", round(mean(sepsis_imputed$anchor_age), 1)),
             color = "red", hjust = 0) +
    labs(title = "Age Distribution of ICU Sepsis Cohort",
         x = "Age (years)", y = "Count") +
    theme_minimal()

# check cleaning: proportion of prolonged stay across pipeline stages
mean(sepsis_with_labs$prolonged_stay)
mean(sepsis_clean$prolonged_stay)
mean(sepsis_imputed$prolonged_stay)
```

# Model 1:

## Logistic Regression: Lasso

```{r}
# 80/20 train/test split
set.seed(42)
train_idx <- sample(1:nrow(X), 0.8 * nrow(X))
X_train <- X[train_idx, ]
X_test  <- X[-train_idx, ]
y_train <- y[train_idx]
y_test  <- y[-train_idx]

# LASSO logistic regression: same as logistic regression but penalizes 
# small coefficients to zero — automatic variable selection
# Ridge: shrinks but keeps all predictors
# Elastic Net: mix of LASSO and Ridge
# LASSO chosen for variable selection despite similar AUC across all models
# Youden threshold: maximizes sensitivity + specificity instead of default 0.5

# LASSO: retrain on training set
set.seed(42)
cv_lasso    <- cv.glmnet(X_train, y_train, family = "binomial", alpha = 1, nfolds = 10)
lasso_model <- glmnet(X_train, y_train, family = "binomial", alpha = 1, lambda = cv_lasso$lambda.min)
lasso_pred  <- as.numeric(predict(lasso_model, newx = X_test, type = "response"))
roc_lasso   <- roc(y_test, lasso_pred)
auc_lasso   <- auc(roc_lasso)

# LASSO: performance on test set at Youden threshold
thresh_lasso <- best_threshold(roc_lasso)
m_lasso      <- calc_metrics(lasso_pred, y_test, thresh_lasso)
cat(sprintf("AUC: %.3f\nThreshold: %.3f\nSensitivity: %.1f%%\nSpecificity: %.1f%%\n\n",
            auc_lasso, thresh_lasso, m_lasso[1], m_lasso[2]))

# LASSO: selected predictors (trained on training set)
coef(lasso_model) |>
  as.matrix() |>
  as.data.frame() |>
  rownames_to_column("Variable") |>
  filter(Variable != "(Intercept)", s0 != 0) |>
  arrange(desc(abs(s0))) |>
  print()


# extended check: all negative coefficient variables vs mortality
sepsis_imputed |>
  left_join(admissions |> select(subject_id, hospital_expire_flag), by = "subject_id") |>
  group_by(hospital_expire_flag) |>
  summarise(
    mean_age       = mean(anchor_age, na.rm = TRUE),
    mean_lactate   = mean(lactate[is.finite(lactate)], na.rm = TRUE),
    mean_resp_rate = mean(resp_rate, na.rm = TRUE),
    mean_dbp       = mean(dbp, na.rm = TRUE),
    mean_sbp       = mean(sbp, na.rm = TRUE),
    mean_los       = mean(los_days, na.rm = TRUE)
  )

# deaths had higher lactate but shorter LOS — explaining negative lactate coefficient
# hospital_expire_flag not included as predictor — mortality occurs during the stay
# and is not known at admission (competing risk, not a baseline variable)

```

## Graphing ROC Curves

```{r}

# build roc_df from test set ROC objects
lab_lr    <- paste0("Logistic Regression\nAUC = ", round(auc_lr, 3))
lab_ridge <- paste0("Ridge\nAUC = ", round(auc_ridge, 3))
lab_lasso <- paste0("LASSO\nAUC = ", round(auc_lasso, 3))
lab_enet  <- paste0("Elastic Net\nAUC = ", round(auc_enet, 3))

roc_df <- bind_rows(
  data.frame(FPR = 1 - roc_lr$specificities,    TPR = roc_lr$sensitivities,    Model = lab_lr),
  data.frame(FPR = 1 - roc_ridge$specificities,  TPR = roc_ridge$sensitivities,  Model = lab_ridge),
  data.frame(FPR = 1 - roc_lasso$specificities,  TPR = roc_lasso$sensitivities,  Model = lab_lasso),
  data.frame(FPR = 1 - roc_enet$specificities,   TPR = roc_enet$sensitivities,   Model = lab_enet)
)
#4 graphs
ggplot(roc_df, aes(x = FPR, y = TPR, color = Model)) +
  geom_line(lwd = 1) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "gray50") +
  facet_wrap(~ Model, ncol = 2) +
  scale_color_manual(values = setNames(
   c("cyan", "blue", "magenta", "green"),
    c(lab_lr, lab_ridge, lab_lasso, lab_enet)
  )) +
  labs(
    title = "ROC Curves by Model",
    x     = "False Positive Rate (1 - Specificity)",
    y     = "True Positive Rate (Sensitivity)"
  ) +
  theme_minimal() +
  theme(
    strip.text      = element_text(face = "bold"),
    legend.position = "none"
  )

# combined all 4 into one graph
ggplot(roc_df, aes(x = FPR, y = TPR, color = Model)) +
  geom_line(lwd = 1) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "gray50") +
  scale_color_manual(values = setNames(
    c("cyan", "blue", "magenta", "green"),
    c(lab_lr, lab_ridge, lab_lasso, lab_enet)
  )) +
  labs(
    title = "ROC Curves by Model",
    x     = "False Positive Rate (1 - Specificity)",
    y     = "True Positive Rate (Sensitivity)"
  ) +
  theme_minimal() +
  theme(legend.position = "bottom")
```
