---
title: "Machine Learning Prediction of Prolonged ICU Stay in Sepsis Patients Using MIMIC-IV"
subtitle: "Biostat 203B Final Project"
author: 
  - name: "Charlotte Lee and 206782165"
  - name: "Yasmine Tani and Brandon Brushwyler"
date: "March 20, 2026"
execute:
  cache: true
format:
  html:
    theme: cosmo
    embed-resources: true
    number-sections: true
    toc: true
    toc-depth: 4
    toc-location: left
    code-fold: false
---

# Abstract

Sepsis is a life-threatening condition and a major cause of Intensive Care Unit (ICU) admission and mortality worldwide. Patients with sepsis who experience prolonged ICU stays (≥14 days) face worse outcomes and increased healthcare costs, yet few studies have developed prediction tools specifically targeting extended length of stay at the time of ICU admission.

This study evaluates six machine learning models for predicting prolonged ICU stay in adult sepsis patients using retrospective data from the MIMIC-IV Clinical Database (n = 4,179). Predictors included first 24-hour worst-case vital signs, laboratory values, demographics, comorbidities derived from ICD-9 and ICD-10 codes, and three clinically motivated interaction terms (CKD×Diabetes, Chronic Liver×CVD, GCS×Vasopressor). Missing values were handled through mean imputation for symmetric variables and median imputation for skewed variables. Models compared included logistic regression, LASSO, Ridge, Elastic Net, Random Forest, and XGBoost, evaluated using AUC and Youden-optimal thresholds on a held-out 20% test set.

XGBoost achieved the highest AUC (0.683), closely followed by Ridge (0.681) and LASSO (0.680). Random Forest performed comparably (AUC = 0.677). Differences across models were minimal (range 0.677– 0.683), suggesting predominantly linear relationships between predictors and prolonged stay. Key predictors included chronic liver disease, cardiovascular disease, Glasgow Coma Scale score, and insurance type. LASSO retained the CKD×Diabetes interaction while dropping diabetes as a main effect, consistent with diabetic nephropathy as the clinically meaningful entity.

These findings suggest that comorbidity burden at admission drives prolonged ICU stay more than acute physiological derangements alone, and that simpler interpretable models perform comparably to complex nonlinear approaches. Given its comparable performance and interpretable coefficients, LASSO was selected for clinical deployment in the accompanying prediction application.

# Introduction

## Background

## Literature Review

## Problem Statement

## Objectives

# Methods

## Data Source

Display machine information:

```{r}
sessionInfo()
```

Display my machine memory.

```{r}
memuse::Sys.meminfo()
```

Load database libraries and the tidyverse frontend:

```{r}
#| label: setup
#| cache: false
#| message: false
#| warning: false
library(bigrquery)
library(dbplyr)
library(DBI)
library(gt)
library(gtsummary)
library(tidyverse)
library(glmnet)
library(pROC)
library(corrplot)
library(tidymodels)
library(ranger)
library(vip)
library(xgboost)
library(styler)

# Style my current file
#styler::style_file("final_project/final.qmd")
```

## Connect to The Data

```{r}
#| cache: false  # overrides the YAML cache: true for this chunk only

bq_auth(path = "biostat-203b-2026-winter-92fefbfab477.json")
con_bq <- dbConnect(
  bigrquery::bigquery(),
  project = "biostat-203b-2025-winter",
  dataset = "mimiciv_3_1",
  billing = "biostat-203b-2025-winter"
)
```

Data were obtained from MIMIC-IV (Medical Information Mart for Intensive Care), a large single-center critical care database containing de-identified electronic health records from Beth Israel Deaconess Medical Center (BIDMC) in Boston, MA. MIMIC-IV contains data from over 50,000 ICU admissions and includes diagnoses, laboratory results, vital signs, medications, and administrative data. Access to MIMIC-IV requires completion of a data use agreement and CITI human subjects training through PhysioNet, ensuring appropriate credentialing and privacy protections. Data were accessed via Google BigQuery using the `bigrquery` and `dbplyr` R packages, enabling SQL-based querying of large tables without loading full datasets into memory. All filtering and aggregation were performed server-side on BigQuery before collecting results, consistent with best practices for handling large-scale clinical databases.

**Inclusion criteria:** Adult patients (age ≥18) with a primary sepsis diagnosis (ICD-9 or ICD-10, sequence number = 1) admitted to the ICU. Only the first qualifying admission per patient was retained to satisfy the independence assumption of logistic regression.

**Exclusion criteria.** Patients who were flagged to have died within 48 hours of ICU admission were excluded to address competing risk bias, these patients cannot by definition have a prolonged stay, and their inclusion would cause high-severity markers to appear falsely protective. 

```{r}
# cohort flow - how many patients at each step
cohort_flow <- tibble(
  Step = c(
    "Adult sepsis ICU admissions",
    "After outcome filter (LOS > 0)",
    "After first admission filter",
    "After 48h early death exclusion",
    "Final analytic cohort"
  ),
  N = c(5017, 4953, 4478, 4179, 4179),
  label = c(
    "5,017",
    "4,953 (-64 same-day discharges)",
    "4,478 (-475 duplicate admissions)",
    "4,179 (-299 early deaths)",
    "4,179 (final)"
  )
) |>
  mutate(Step = factor(Step, levels = rev(Step)))

ggplot(cohort_flow, aes(x = N, y = Step)) +
  geom_col(fill = "steelblue", width = 0.5) +
  geom_text(aes(label = label), hjust = -0.05, size = 3.5) +
  labs(title = "Cohort Construction Flow",
       x = "Number of Patients", y = NULL) +
  xlim(0, 9000) +   # increased to fit long labels
  theme_minimal()
```

## Variables and Features

**Outcome:** Prolonged ICU stay, defined as hospital length of stay ≥14 days (binary: 1 = prolonged, 0 = normal).

**Predictors (19 total):**

**Vital signs (4):** Maximum heart rate, minimum mean arterial pressure (MAP), maximum respiratory rate, and maximum temperature during the first 24 hours of ICU admission. Worst-case values were used consistent with severity scoring conventions (APACHE II, SOFA). MAP was selected over systolic and diastolic blood pressure due to its use as the primary hemodynamic target in the Surviving Sepsis Campaign (MAP \>65 mmHg) and its inclusion in SOFA scoring. SBP and DBP were excluded as multicollinear with MAP.

**Laboratory values (3):** Maximum creatinine, maximum white blood cell count (WBC), and minimum platelets during **the first 24 hours. Bilirubin and lactate were excluded after sensitivity analysis showed no change in predictive performance (AUC 0.680 with and without), and both had \>15% missingness with patterns consistent with missing-not-at-random in ICU settings.**

**Neurological (1):** Minimum Glasgow Coma Scale (GCS) total score during the first 24 hours, capturing neurological severity not represented by other predictors.

**Demographics (3):** Age, gender (binary: male = 1), and race (collapsed to White, Black, Hispanic, Asian, Other, Unknown). \*Social determinants (1):\* Insurance type (Medicare, Medicaid, Private, Other, Unknown), which showed a 12% difference in prolonged stay prevalence across categories.

**Comorbidities (7)**: Hypertension, diabetes, chronic kidney disease (CKD), cardiovascular disease (CVD), chronic liver disease, immunosuppression, and vasopressor use — all derived from ICD-9 and ICD-10 secondary diagnosis codes or medication administration records.

**Interaction terms (3):** CKD×Diabetes (diabetic nephropathy), Chronic Liver×CVD (cardiohepatic syndrome), and GCS×Vasopressor (combined neurological and hemodynamic instability). These were included based on clinical reasoning that the joint presence of these conditions represents syndromes with distinct prognostic implications beyond their main effects.

**Pre-processing:** Outlier removal was performed by setting physiologically impossible values to missing (NA). Bounds were defined at biological limits — not clinical abnormality thresholds — to remove data entry errors while preserving true disease severity. Specific bounds applied were: heart rate (\<20 or \>300 bpm), mean arterial pressure (≤0 or \>200 mmHg), respiratory rate (\<5 or \>80 breaths/min), temperature (\<25°C or \>45°C), lactate (\>30 mmol/L), white blood cell count (\>500 K/µL), platelets (\>3,000 K/µL), creatinine (\>50 mg/dL), and Glasgow Coma Scale total (\<3 or \>15, as the scale is defined over this range by construction). Insurance type did not require outlier cleaning as it is categorical with fixed levels. Extreme values that were retained — such as respiratory rate of 73, WBC of 267, and temperature of 42.5°C — reflect physiologically possible values in severely ill sepsis patients. Following outlier removal, remaining missing values were imputed using the column median computed from the full cohort.

### Sepsis Diagnosis and Adult Filter

```{r}
# load raw tables
diagnoses_icd_tbl <- tbl(con_bq, "diagnoses_icd") |>
  arrange(subject_id, hadm_id)

d_icd_diagnoses_tbl <- tbl(con_bq, "d_icd_diagnoses")

patients_tbl <- tbl(con_bq, "patients") |>
  select(subject_id, anchor_age, gender)

# match sepsis icd codes from lookup
sepsis_codes_tbl <- d_icd_diagnoses_tbl |>
  filter(sql("LOWER(long_title) LIKE '%sepsis%'")) |>
  select(icd_code, icd_version, long_title)

sepsis_codes_tbl |> collect() |> print(n = 50)

# keep only primary diagnosis (seq_num == 1)
sepsis_cohort_tbl <- diagnoses_icd_tbl |>
  filter(seq_num == 1) |>
  inner_join(sepsis_codes_tbl, by = c("icd_code", "icd_version")) |>
  select(subject_id, hadm_id, seq_num, icd_code, icd_version, long_title)

# filter adults only
adults_tbl <- patients_tbl |>
  filter(anchor_age >= 18)

sepsis_adult_cohort_tbl <- sepsis_cohort_tbl |>
  inner_join(adults_tbl, by = "subject_id")

cohort_df <- sepsis_adult_cohort_tbl |> collect()

```

### ICU Patients

```{r}
# restrict to ICU patients only upfront
icu_hadm_ids <- tbl(con_bq, "icustays") |>
  select(hadm_id) |>
  distinct() |>
  collect() |>
  pull(hadm_id)

cohort_df <- cohort_df |>
  filter(hadm_id %in% icu_hadm_ids)

# pulls LOS + race + insurance in one query instead of two
admissions_tbl <- tbl(con_bq, "admissions") |>
  select(subject_id, hadm_id, admittime, dischtime, race, insurance) |>
  mutate(los_days = sql("TIMESTAMP_DIFF(dischtime, admittime, DAY)")) |>
  collect()

# join onto cohort and create outcome variable (prolonged stay >= 14 days)
sepsis_with_outcome <- cohort_df |>
  inner_join(admissions_tbl, by = c("subject_id", "hadm_id")) |>
  filter(!is.na(los_days), los_days > 0) |>
  mutate(prolonged_stay = if_else(los_days >= 14, 1L, 0L))
```

### Comorbidities

```{r}
cohort_ids <- sepsis_with_demo$hadm_id

# pull secondary diagnoses — both ICD-9 AND ICD-10 to capture all admissions
# collect AFTER all BQ-side filters
secondary_dx <- tbl(con_bq, "diagnoses_icd") |>
  filter(
    seq_num > 1,
    hadm_id %in% !!cohort_ids
  ) |>
  select(subject_id, hadm_id, icd_code, icd_version) |>
  collect()

comorbidity_flags <- secondary_dx |>
  mutate(
    # hypertension — ICD-10: I10 | ICD-9: 401-405
    hypertension = if_else(
      (icd_version == 10 & startsWith(icd_code, "I10")) |
      (icd_version == 9  & (startsWith(icd_code, "401") | startsWith(icd_code, "402") |
                             startsWith(icd_code, "403") | startsWith(icd_code, "404") |
                             startsWith(icd_code, "405"))),
      1L, 0L),

    # diabetes — ICD-10: E10/E11 | ICD-9: 250
    diabetes = if_else(
      (icd_version == 10 & (startsWith(icd_code, "E10") | startsWith(icd_code, "E11"))) |
      (icd_version == 9  &  startsWith(icd_code, "250")),
      1L, 0L),

    # chronic kidney disease — ICD-10: N18 | ICD-9: 585-586
    ckd = if_else(
      (icd_version == 10 & startsWith(icd_code, "N18")) |
      (icd_version == 9  & (startsWith(icd_code, "585") | startsWith(icd_code, "586"))),
      1L, 0L),

    # cardiovascular disease — ICD-10: I20/I21/I25/I48/I50/I70/I71
    #                           ICD-9: 410-414/427/428/440/441
    cvd = if_else(
      (icd_version == 10 & (
        startsWith(icd_code, "I20") | startsWith(icd_code, "I21") |
        startsWith(icd_code, "I25") | startsWith(icd_code, "I48") |
        startsWith(icd_code, "I50") | startsWith(icd_code, "I70") |
        startsWith(icd_code, "I71"))) |
      (icd_version == 9  & (
        startsWith(icd_code, "410") | startsWith(icd_code, "411") |
        startsWith(icd_code, "412") | startsWith(icd_code, "413") |
        startsWith(icd_code, "414") | startsWith(icd_code, "427") |
        startsWith(icd_code, "428") | startsWith(icd_code, "440") |
        startsWith(icd_code, "441"))),
      1L, 0L),

    # chronic liver disease — ICD-10: K70/K73/K74 | ICD-9: 571
    chronic_liver = if_else(
      (icd_version == 10 & (
        startsWith(icd_code, "K70") | startsWith(icd_code, "K73") |
        startsWith(icd_code, "K74"))) |
      (icd_version == 9  &  startsWith(icd_code, "571")),
      1L, 0L),

    # immunosuppression — ICD-10: D80/D81/D83/Z79/Z94/C81-C83/C91/C92
    #                     ICD-9: 279/V42/V45/200-202/204/205
    immunosuppression = if_else(
      (icd_version == 10 & (
        startsWith(icd_code, "D80") | startsWith(icd_code, "D81") |
        startsWith(icd_code, "D83") | startsWith(icd_code, "Z79") |
        startsWith(icd_code, "Z94") | startsWith(icd_code, "C81") |
        startsWith(icd_code, "C82") | startsWith(icd_code, "C83") |
        startsWith(icd_code, "C91") | startsWith(icd_code, "C92"))) |
      (icd_version == 9  & (
        startsWith(icd_code, "279") | startsWith(icd_code, "V42") |
        startsWith(icd_code, "V45") | startsWith(icd_code, "200") |
        startsWith(icd_code, "201") | startsWith(icd_code, "202") |
        startsWith(icd_code, "204") | startsWith(icd_code, "205"))),
      1L, 0L)
  ) |>
  group_by(subject_id, hadm_id) |>
  summarise(across(hypertension:immunosuppression, max), .groups = "drop") |>
  mutate(across(hypertension:immunosuppression, ~replace_na(.x, 0L)))

sepsis_with_comorbidities <- sepsis_with_demo |>
  left_join(comorbidity_flags, by = c("subject_id", "hadm_id")) |>
  mutate(across(hypertension:immunosuppression, ~replace_na(.x, 0L)))

# check 
cat("Rows after joining comorbidities: ", nrow(sepsis_with_comorbidities), "\n")
comorbidity_flags |>
  summarise(across(hypertension:immunosuppression, ~round(mean(.x)*100,1))) |>
  print(width=Inf)
dim(sepsis_with_comorbidities)
```

### ICU admission Times

```{r}
### 3.3 ICU Admission Times (for 24h vital/lab windows)

cohort_ids <- sepsis_with_comorbidities$hadm_id

# get earliest ICU admission time per stay — used to define 24h window below
icu_admissions_bq <- tbl(con_bq, "icustays") |>
  filter(hadm_id %in% !!cohort_ids) |>
  group_by(subject_id, hadm_id) |>
  summarise(intime = min(intime, na.rm = TRUE), .groups = "drop")

icustays_raw <- icu_admissions_bq |> collect()
icu_admissions <- icustays_raw

cat("ICU stays pulled:", nrow(icustays_raw), "\n")
```

### Vital Items

```{r}
# pull vitals from chartevents within first 24h of ICU admission
# itemid 223761 is fahrenheit so converting to celsius
vitals_raw <- tbl(con_bq, "chartevents") |>
  filter(
    itemid %in% c(220045, 220179, 220050, 220180, 220051,
                  220052, 220181, 220210, 223762, 223761),
    hadm_id %in% !!cohort_ids,
    !is.na(valuenum)
  ) |>
  select(subject_id, hadm_id, itemid, charttime, valuenum) |>
  inner_join(icu_admissions_bq, by = c("subject_id", "hadm_id")) |>
  filter(
    charttime >= intime,
    charttime <= sql("TIMESTAMP_ADD(intime, INTERVAL 24 HOUR)")
  ) |>
  mutate(
    valuenum = if_else(itemid == 223761, (valuenum - 32) * 5/9, valuenum)
  ) |>
  collect()   # now collects only 24h window rows, not the whole table

# map itemids to vital names (some vitals have multiple itemids)
vital_items <- tibble::tibble(
  itemid     = c(220045, 220179, 220050, 220180, 220051,
                 220052, 220181, 220210, 223762, 223761),
  vital_name = c("heart_rate", "sbp",  "sbp",  "dbp",  "dbp",
                 "map",  "map",  "resp_rate", "temperature", "temperature")
)
  #map combined then min of both found which is clincially standard approach

# summarise each vital per admission (worst values)
vitals_summary <- vitals_raw |>
  left_join(vital_items, by = "itemid") |>
  group_by(subject_id, hadm_id, vital_name) |>
  summarise(
    value = case_when(
      # MAX — tachycardia indicates cardiac stress/infection severity
      first(vital_name) == "heart_rate"  ~ max(valuenum),
      
      # MIN — hypotension indicates shock/hemodynamic instability
      first(vital_name) == "sbp"         ~ min(valuenum),
      first(vital_name) == "dbp"         ~ min(valuenum),
      first(vital_name) == "map"         ~ min(valuenum),
      
      # MAX — tachypnea indicates respiratory distress/metabolic acidosis
      first(vital_name) == "resp_rate"   ~ max(valuenum),
      
      # MAX — peak fever indicates infection severity
      # (hypothermia in sepsis also dangerous but less common)
      first(vital_name) == "temperature" ~ max(valuenum),
    ),
    .groups = "drop"
  ) |>
  pivot_wider(names_from = vital_name, values_from = value)
```

### Lab Events

```{r}
# pull labs from labevents within first 24h of ICU admission
# creatinine (50912), WBC (51301), platelets (51265) only
# bilirubin and lactate excluded after sensitivity analysis showed
# no change in predictive performance (AUC 0.680 with and without)
labs_raw <- tbl(con_bq, "labevents") |>
  filter(
    itemid %in% c(50912, 51301, 51265),
    hadm_id %in% !!cohort_ids,
    !is.na(valuenum)
  ) |>
  select(subject_id, hadm_id, itemid, charttime, valuenum) |>
  inner_join(icu_admissions_bq, by = c("subject_id", "hadm_id")) |>
  filter(
    charttime >= intime,
    charttime <= sql("TIMESTAMP_ADD(intime, INTERVAL 24 HOUR)")
  ) |>
  collect()

lab_items <- tibble::tibble(
  itemid   = c(50912,        51301,  51265),
  lab_name = c("creatinine", "wbc",  "platelets")
)

labs_summary <- labs_raw |>
  left_join(lab_items, by = "itemid") |>
  group_by(subject_id, hadm_id, lab_name) |>
  summarise(
    value = case_when(
      first(lab_name) == "creatinine" ~ max(valuenum),
      first(lab_name) == "wbc"        ~ max(valuenum),
      first(lab_name) == "platelets"  ~ min(valuenum)
    ),
    .groups = "drop"
  ) |>
  pivot_wider(names_from = lab_name, values_from = value)

cat("Rows after joining labs:", nrow(labs_summary), "\n")
```

### Glasgow Coma Scale (first 24 hr minimum)

```{r}
### 3.6 GCS (first 24h minimum — lowest score = worst neurological state)
gcs_raw <- tbl(con_bq, "chartevents") |>
  filter(
    itemid %in% c(220739, 223900, 223901),
    hadm_id %in% !!cohort_ids,
    !is.na(valuenum)
  ) |>
  select(subject_id, hadm_id, itemid, charttime, valuenum) |>
  inner_join(icu_admissions_bq, by = c("subject_id", "hadm_id")) |>
  filter(
    charttime >= intime,
    charttime <= sql("TIMESTAMP_ADD(intime, INTERVAL 24 HOUR)")
  ) |>
  collect() |>
  group_by(subject_id, hadm_id, itemid) |>
  summarise(value = min(valuenum, na.rm = TRUE), .groups = "drop") |>
  pivot_wider(names_from = itemid, values_from = value) |>
  rename(gcs_eye    = `220739`,
         gcs_verbal = `223900`,
         gcs_motor  = `223901`) |>
  mutate(gcs_total = gcs_eye + gcs_verbal + gcs_motor) |>
  select(subject_id, hadm_id, gcs_total)

cat("Patients with GCS data:", nrow(gcs_raw), "\n")
summary(gcs_raw$gcs_total)
```

### Vasopressors (first 24h binary flags)

```{r}
# vasopressor itemids — norepinephrine, epinephrine, dopamine, vasopressin, phenylephrine
vasopressor_ids <- c(221906, 221289, 221662, 222315, 221749)

vasopressors_raw <- tbl(con_bq, "inputevents") |>
  filter(
    itemid %in% vasopressor_ids,
    hadm_id %in% !!cohort_ids
  ) |>
  select(subject_id, hadm_id, itemid, starttime) |>
  inner_join(icu_admissions_bq, by = c("subject_id", "hadm_id")) |>
  filter(
    starttime >= intime,
    starttime <= sql("TIMESTAMP_ADD(intime, INTERVAL 24 HOUR)")
  ) |>
  collect()

# binary flag — 1 if patient received vasopressors in first 24h, 0 otherwise
vasopressor_flags <- vasopressors_raw |>
  group_by(subject_id, hadm_id) |>
  summarise(vasopressor = 1L, .groups = "drop")

# check
cat("Patients with vasopressors in first 24h:", nrow(vasopressor_flags), "\n")
```

### Joining all Predictors

```{r}
# Safety Check + Joining all Predictors

# ICU filter already applied in section 1 — this is just a safety check
sepsis_icu <- sepsis_with_comorbidities

sepsis_with_vitals <- sepsis_icu |>
  left_join(vitals_summary, by = c("subject_id", "hadm_id"))

sepsis_with_labs <- sepsis_with_vitals |>
  left_join(labs_summary,      by = c("subject_id", "hadm_id")) |>
  left_join(vasopressor_flags, by = c("subject_id", "hadm_id")) |>
  left_join(gcs_raw,           by = c("subject_id", "hadm_id")) |>  # added GCS
  mutate(vasopressor = replace_na(vasopressor, 0L)) |>
  select(-sbp, -dbp)   # dropped sbp monitor 1, sdp monitor 2, dbp (both monitors), sbpmulticollinear variables

cat("Final rows:", nrow(sepsis_with_labs), "\n")
cat("Vasopressor use (%):", round(mean(sepsis_with_labs$vasopressor)*100, 1), "\n")
cat("GCS missing (%):", round(mean(is.na(sepsis_with_labs$gcs_total))*100, 1), "\n")
```

### Check Raw Data

```{r}
#raw data view
#summary(sepsis_with_labs)

# check
cat("raw final dataset:", nrow(sepsis_icu), "\n")
cat("Final rows:", nrow(sepsis_with_labs), "\n")

# check % missingness across all columns
sepsis_with_labs %>%
  summarise(across(everything(), ~ round(mean(is.na(.x)) * 100, 1))) %>%
  print(width = Inf)
```

### Outlier Cleaning

```{r}
# Outlier cleaning note:
# bounds set at physiologically IMPOSSIBLE limits, not clinical abnormality limits
# goal is to remove data entry errors while preserving true disease severity
# MAP lower bound set to >0 (not >=0) since zero BP = death/error
# resp_rate max 73, wbc max 267 retained as physiologically 
# possible extreme values in severely ill sepsis patients
# temperature max 42.5C retained as real severe fever


# remove physiologically IMPOSSIBLE values only — not clinical abnormalities
# goal: remove data entry errors, NOT filter out severely ill patients
# a lactate of 28 is real and kept; a lactate of 1,276,103 is a data error
sepsis_clean <- sepsis_with_labs |>
  mutate(
    # heart rate: no human survives >300, <20 is incompatible with life
    heart_rate  = if_else(heart_rate > 300 | heart_rate < 20,
                          NA_real_, heart_rate),

    # map: zero or negative impossible, >200 is data error
    map         = if_else(map <= 0 | map > 200, NA_real_, map),

    # resp rate: <5 impossible, >80 is data error
    resp_rate   = if_else(resp_rate < 5 | resp_rate > 80,
                          NA_real_, resp_rate),

    # temperature: <25 incompatible with life, >45 lethal/data error
    temperature = if_else(temperature < 25 | temperature > 45,
                          NA_real_, temperature),

    # wbc: >500 is beyond any recorded human value
    wbc         = if_else(wbc > 500, NA_real_, wbc),

    # platelets: >3000 is beyond any recorded human value
    platelets   = if_else(platelets > 3000, NA_real_, platelets),

    # creatinine: >50 is beyond any recorded human value
    creatinine  = if_else(creatinine > 50, NA_real_, creatinine),

    # GCS: scale is 3-15 by definition, outside = data error
    gcs_total   = if_else(gcs_total < 3 | gcs_total > 15,
                          NA_real_, gcs_total)
  )

# insurance doesn't need outlier cleaning — categorical with fixed levels

# check missingness after cleaning
sepsis_clean |>
  summarise(across(everything(), ~ round(mean(is.na(.x)) * 100, 1))) |>
  print(width = Inf)

# verify crazy values are gone
summary(sepsis_clean[, c("resp_rate", "temperature",
                          "map", "wbc", "creatinine", "gcs_total")])

```

### Imputation

```{r}
# step 1: keep first admission per patient
sepsis_clean <- sepsis_clean |>
  arrange(subject_id, admittime) |>
  group_by(subject_id) |>
  slice(1) |>
  ungroup()

cat("After first admission filter:", nrow(sepsis_clean), "\n")
cat("Duplicate subject_ids:", sum(duplicated(sepsis_clean$subject_id)), "\n")

# step 2: remove patients who died within 48h of ICU admission
# these patients cannot have prolonged stay by definition
# and create competing risk bias — sickest patients die quickly,
# making high lactate/vasopressor falsely appear protective
early_deaths <- tbl(con_bq, "admissions") |>
  select(subject_id, hadm_id, hospital_expire_flag, deathtime) |>
  collect()

sepsis_clean <- sepsis_clean |>
  left_join(early_deaths, by = c("subject_id", "hadm_id")) |>
  left_join(icustays_raw |> select(subject_id, hadm_id, intime),
            by = c("subject_id", "hadm_id")) |>
  filter(
    hospital_expire_flag == 0 |
    is.na(deathtime) |
    as.numeric(difftime(deathtime, intime, units = "hours")) > 48
  ) |>
  select(-hospital_expire_flag, -deathtime, -intime)

cat("After removing early deaths:", nrow(sepsis_clean), "\n")

# step 3: fix Inf/-Inf then median impute remaining NAs
# Inf arises when max/min was called on a patient with ALL NA values in 24h window
# median chosen over mean due to right-skewed distributions in clinical variables
# gcs_total: low score = worst neurological state, median impute missing values
sepsis_imputed <- sepsis_clean |>
  mutate(across(
    .cols = c(heart_rate, map, resp_rate, temperature,
              creatinine, bilirubin, wbc, platelets, lactate, gcs_total),
    ~ if_else(is.infinite(.x), NA_real_, .x)
  )) |>
  mutate(across(
    .cols = c(heart_rate, map, resp_rate, temperature,
              creatinine, bilirubin, wbc, platelets, lactate, gcs_total),
    ~ coalesce(.x, median(.x, na.rm = TRUE))
  ))

# confirm no remaining missingness
sepsis_imputed |>
  summarise(across(c(heart_rate, map, resp_rate, temperature,
                     creatinine, bilirubin, wbc, platelets, 
                     lactate, gcs_total),
                   ~ round(median(is.na(.x)) * 100, 1))) |>
  print()

cat("Final patients:", nrow(sepsis_imputed), "\n")
cat("Duplicate subject_ids:", sum(duplicated(sepsis_imputed$subject_id)), "\n")

```

## Analytic Approach and Modeling

**Index time:** ICU admission. The prediction window was the first 24 hours of ICU admission, using worst-case values to capture peak physiological derangement.

**Models:** Six classification models were trained and compared:

1.  *Logistic Regression* — baseline linear model
2.  *LASSO* (α = 1) — L1 penalization for automatic variable selection
3.  *Ridge* (α = 0) — L2 penalization retaining all predictors
4.  *Elastic Net* (α = 0.5) — combination of L1 and L2 penalties
5.  *Random Forest* — ensemble of decision trees with bootstrap aggregation
6.  *XGBoost* — gradient boosted trees with sequential error correction

All models were implemented in R using the `glmnet` package for regularized regression and the `tidymodels` framework for Random Forest (`ranger`) and XGBoost (`xgboost`).

## Model Matrix

```{r}
# race included as categorical predictor — dummy coded automatically by model.matrix
# gender recoded as binary (1 = M, 0 = F)
# vasopressor: binary flag — received any vasopressor in first 24h of ICU admission
# model matrix
#insurance: captures social determinants of health — meaningful variation in prolonged
#            stay across categories (Other 46.1% vs Medicare 34.1%)
model_vars <- c("anchor_age", "gender", "race_collapsed",
                "heart_rate", "map", "resp_rate",
                "temperature", "creatinine",
                "wbc", "platelets", "gcs_total",
                "hypertension", "diabetes", "ckd", "cvd",
                "chronic_liver", "immunosuppression",
                "vasopressor", "insurance")

model_df <- sepsis_imputed |>
  mutate(
    gender = if_else(gender == "M", 1L, 0L),
    race_collapsed = case_when(
      str_detect(race, "WHITE")    ~ "WHITE",
      str_detect(race, "BLACK")    ~ "BLACK",
      str_detect(race, "HISPANIC") ~ "HISPANIC",
      str_detect(race, "ASIAN")    ~ "ASIAN",
      race %in% c("UNKNOWN", "UNABLE TO OBTAIN",
                  "PATIENT DECLINED TO ANSWER") ~ "UNKNOWN",
      TRUE ~ "OTHER"
    ) |> as.factor(),
    insurance = if_else(is.na(insurance), "Unknown",
                        as.character(insurance)) |> as.factor()
  ) |>
  select(all_of(model_vars), prolonged_stay) |>
  mutate(across(where(is.numeric), ~if_else(is.infinite(.x), NA_real_, .x))) |>
  drop_na()

cat("Rows in model_df:", nrow(model_df), "\n")
cat("Total rows in sepsis_imputed:", nrow(sepsis_imputed), "\n")
cat("Rows dropped by drop_na():", nrow(sepsis_imputed) - nrow(model_df), "\n")
cat("Insurance distribution:\n")
table(model_df$insurance) |> print()
cat("Outcome distribution:\n")
prop.table(table(model_df$prolonged_stay)) |> round(3) |> print()

X <- model.matrix(~ . - prolonged_stay - 1, data = model_df)
y <- model_df$prolonged_stay

cat("Model matrix dimensions:", dim(X), "\n")
cat("Number of predictors:", ncol(X), "\n")

best_threshold <- function(roc_obj) {
  idx <- which.max(roc_obj$sensitivities + roc_obj$specificities - 1)
  roc_obj$thresholds[idx]
}

calc_metrics <- function(pred_prob, actual, threshold) {
  pred_class <- if_else(pred_prob >= threshold, 1L, 0L)
  sens <- mean(pred_class[actual == 1] == 1) * 100
  spec <- mean(pred_class[actual == 0] == 0) * 100
  c(sens, spec)
}

set.seed(42)
train_idx <- sample(1:nrow(X), 0.8 * nrow(X))
X_train <- X[train_idx, ]
X_test  <- X[-train_idx, ]
y_train <- y[train_idx]
y_test  <- y[-train_idx]

# add interaction terms to X:
X_interact <- model.matrix(
  ~ . - prolonged_stay - 1 + 
    chronic_liver:cvd +      # liver disease + cardiovascular disease
    ckd:diabetes +           # diabetic nephropathy
    gcs_total:vasopressor,   # neurological impairment + hemodynamic instability
  data = model_df
)

# use X_interact instead of X for all models
X_train_i <- X_interact[train_idx, ]
X_test_i  <- X_interact[-train_idx, ]
```

### Logistic Regression (Simple, LASSO, Ridge, Elastic Net)

```{r}
# LASSO logistic regression: same as logistic regression but penalizes 
# small coefficients to zero — automatic variable selection
# Ridge: shrinks but keeps all predictors
# Elastic Net: mix of LASSO and Ridge
# LASSO chosen for variable selection despite similar AUC across all models

# simple logistic regression (no regulization)
set.seed(42)
lr_model <- glm(y_train ~ ., data = as.data.frame(X_train_i), family = "binomial")
lr_pred  <- predict(lr_model, newdata = as.data.frame(X_test_i), type = "response")

# LASSO
set.seed(42)
cv_lasso    <- cv.glmnet(X_train_i, y_train, family = "binomial", alpha = 1, nfolds = 10)
lasso_model <- glmnet(X_train_i, y_train, family = "binomial", alpha = 1, lambda = cv_lasso$lambda.min)
lasso_pred  <- as.numeric(predict(lasso_model, newx = X_test_i, type = "response"))

# ridge
set.seed(42)
cv_ridge    <- cv.glmnet(X_train_i, y_train, family = "binomial", alpha = 0, nfolds = 10)
ridge_model <- glmnet(X_train_i, y_train, family = "binomial", alpha = 0, lambda = cv_ridge$lambda.min)
ridge_pred  <- as.numeric(predict(ridge_model, newx = X_test_i, type = "response"))

# elastic net
set.seed(42)
cv_enet    <- cv.glmnet(X_train_i, y_train, family = "binomial", alpha = 0.5, nfolds = 10)
enet_model <- glmnet(X_train_i, y_train, family = "binomial", alpha = 0.5, lambda = cv_enet$lambda.min)
enet_pred  <- as.numeric(predict(enet_model, newx = X_test_i, type = "response"))

# compute ROC objects once
roc_lr    <- suppressMessages(roc(y_test, lr_pred))
roc_lasso <- suppressMessages(roc(y_test, lasso_pred))
roc_ridge <- suppressMessages(roc(y_test, ridge_pred))
roc_enet  <- suppressMessages(roc(y_test, enet_pred))

auc_lr    <- auc(roc_lr)
auc_lasso <- auc(roc_lasso)
auc_ridge <- auc(roc_ridge)
auc_enet  <- auc(roc_enet)

thresh_lr    <- best_threshold(roc_lr)
thresh_lasso <- best_threshold(roc_lasso)
thresh_ridge <- best_threshold(roc_ridge)
thresh_enet  <- best_threshold(roc_enet)

m_lr    <- calc_metrics(lr_pred,    y_test, thresh_lr)
m_lasso <- calc_metrics(lasso_pred, y_test, thresh_lasso)
m_ridge <- calc_metrics(ridge_pred, y_test, thresh_ridge)
m_enet  <- calc_metrics(enet_pred,  y_test, thresh_enet)

# clean results table
results_df <- tibble(
  Model       = c("Logistic Regression", "LASSO", "Ridge", "Elastic Net"),
  AUC         = round(c(auc_lr, auc_lasso, auc_ridge, auc_enet), 3),
  Threshold   = round(c(thresh_lr, thresh_lasso, thresh_ridge, thresh_enet), 3),
  Sensitivity = round(c(m_lr[1], m_lasso[1], m_ridge[1], m_enet[1]), 1),
  Specificity = round(c(m_lr[2], m_lasso[2], m_ridge[2], m_enet[2]), 1)
)
results_df |> gt()

# LASSO selected predictors
lasso_coefs <- coef(lasso_model) |>
  as.matrix() |>
  as.data.frame() |>
  rownames_to_column("Variable") |>
  filter(Variable != "(Intercept)") |>
  mutate(dropped = if_else(s0 == 0, "DROPPED", "kept"))

cat("Dropped by LASSO:\n")
lasso_coefs |> filter(dropped == "DROPPED") |> pull(Variable) |> print()

cat("\nKept by LASSO:\n") 
lasso_coefs |> filter(dropped == "kept")|> arrange(desc(abs(s0))) |> pull(Variable) |> print()
```

### Random Forest for Classification Model

```{r}
# ── Random Forest ─────────────────────────────────────────────────────────────
set.seed(42)
sepsis_split <- initial_split(
  model_df |> mutate(prolonged_stay = as.factor(prolonged_stay)),
  prop = 0.8,
  strata = "prolonged_stay"
)
sepsis_other <- training(sepsis_split)
sepsis_test  <- testing(sepsis_split)

rf_recipe <- recipe(prolonged_stay ~ ., data = sepsis_other) |>
  step_dummy(all_nominal_predictors()) |>
  step_zv(all_numeric_predictors())

rf_mod <- rand_forest(
  mode  = "classification",
  mtry  = tune(),
  trees = tune()
) |>
  set_engine("ranger", importance = "impurity")

rf_wf <- workflow() |>
  add_recipe(rf_recipe) |>
  add_model(rf_mod)

# updated: trees expanded to 200-500 (plot showed upward trend at 300)
# mtry 1-5 unchanged — best mtry=3 was interior to range
param_grid_rf <- grid_regular(
  trees(range = c(200L, 500L)),
  mtry(range  = c(1L, 5L)),
  levels = c(4, 5)
)

set.seed(42)
folds <- vfold_cv(sepsis_other, v = 5, strata = "prolonged_stay")

rf_fit <- rf_wf |>
  tune_grid(
    resamples = folds,
    grid      = param_grid_rf,
    metrics   = metric_set(roc_auc, accuracy)
  )

# diagnosis plot
rf_fit |>
  collect_metrics() |>
  filter(.metric == "roc_auc") |>
  mutate(mtry = as.factor(mtry)) |>
  ggplot(aes(x = trees, y = mean, color = mtry, group = mtry)) +
  geom_line() + geom_point() +
  geom_errorbar(aes(ymin=mean-std_err, ymax=mean+std_err), width=10) +
  labs(x = "Number of Trees", y = "CV AUC",
       title = "Random Forest CV Tuning",
       subtitle = "Trees range expanded to 200-500 based on diagnostic plot") +
  theme_minimal()

rf_fit |> show_best(metric = "roc_auc")

best_rf      <- rf_fit |> select_best(metric = "roc_auc")
final_rf_wf  <- rf_wf  |> finalize_workflow(best_rf)
final_rf_fit <- final_rf_wf |> last_fit(sepsis_split)

final_rf_fit |> collect_metrics()

final_rf_fit |>
  extract_fit_parsnip() |>
  vip(num_features = 19) +
  labs(title = "Random Forest Variable Importance")

final_rf_fit |>
  collect_predictions() |>
  roc_curve(prolonged_stay, .pred_1) |>
  autoplot()

rf_auc <- final_rf_fit |>
  collect_metrics() |>
  filter(.metric == "roc_auc") |>
  pull(.estimate)

cat("Random Forest Test AUC:", round(rf_auc, 3), "\n")

```

### XG Boost Model

```{r}
# ── XGBoost ───────────────────────────────────────────────────────────────────
gb_recipe <- recipe(prolonged_stay ~ ., data = sepsis_other) |>
  step_dummy(all_nominal_predictors()) |>
  step_zv(all_numeric_predictors())

gb_mod <- boost_tree(
  mode       = "classification",
  trees      = 1000,
  tree_depth = tune(),
  learn_rate = tune()
) |>
  set_engine("xgboost")

gb_wf <- workflow() |>
  add_recipe(gb_recipe) |>
  add_model(gb_mod)

# updated: learn_rate narrowed from (-5, 2) to (-5, 0)
# diagnostic plot showed AUC crashes after learn_rate > 1
# concentrates all grid points in the effective region (1e-05 to 1)
param_grid_gb <- grid_regular(
  tree_depth(range = c(1L, 3L)),
  learn_rate(range = c(-5, 0), trans = log10_trans()),
  levels = c(3, 10)
)

set.seed(42)
folds <- vfold_cv(sepsis_other, v = 5, strata = "prolonged_stay")

gb_fit <- gb_wf |>
  tune_grid(
    resamples = folds,
    grid      = param_grid_gb,
    metrics   = metric_set(roc_auc, accuracy)
  )

# diagnosis plot
gb_fit |>
  collect_metrics() |>
  filter(.metric == "roc_auc") |>
  mutate(tree_depth = as.factor(tree_depth)) |>
  ggplot(aes(x = learn_rate, y = mean, color = tree_depth)) +
  geom_point() + geom_line() +
  scale_x_log10() +
  geom_errorbar(aes(ymin=mean-std_err, ymax=mean+std_err), width=0.1) +
  labs(x = "Learning Rate (log scale)", y = "CV AUC",
       title = "XGBoost CV Tuning",
       subtitle = "Learn rate narrowed to 1e-05 to 1 based on diagnostic plot") +
  theme_minimal()

gb_fit |> show_best(metric = "roc_auc")

best_gb      <- gb_fit |> select_best(metric = "roc_auc")
final_gb_wf  <- gb_wf  |> finalize_workflow(best_gb)
final_gb_fit <- final_gb_wf |> last_fit(sepsis_split)

final_gb_fit |> collect_metrics()

final_gb_fit |>
  extract_fit_parsnip() |>
  vip(num_features = 19) +
  labs(title = "XGBoost Variable Importance")

final_gb_fit |>
  collect_predictions() |>
  roc_curve(prolonged_stay, .pred_1) |>
  autoplot()

gb_auc <- final_gb_fit |>
  collect_metrics() |>
  filter(.metric == "roc_auc") |>
  pull(.estimate)

cat("XGBoost Test AUC:", round(gb_auc, 3), "\n")
```

## 4.4 Model Evaluation

**Train/test split:** An 80/20 stratified split was used (n = 3,343 training, n = 836 test), stratified on the outcome to preserve class balance.

**Cross-validation:** 5-fold cross-validation was used for hyperparameter tuning of Random Forest (mtry, trees) and XGBoost (tree_depth, learn_rate).

**Hyperparameter tuning:** Initial tuning ranges were informed by diagnostic plots of CV AUC against hyperparameter values. For Random Forest, the trees range was expanded from 100–300 to 200–500 after diagnostic plots showed an upward trend at the boundary. For XGBoost, the learning rate range was narrowed from (1e-5, 100) to (1e-5, 1) after diagnostic plots revealed AUC collapsed for learn_rate \> 1, concentrating grid points in the effective region.

```{r}
# RF Diagnosis Plot — confirms trees range was appropriate
rf_fit |>
  collect_metrics() |>
  filter(.metric == "roc_auc") |>
  mutate(mtry = as.factor(mtry)) |>
  ggplot(aes(x=trees, y=mean, color=mtry, group=mtry)) +
  geom_line() + geom_point() +
  geom_errorbar(aes(ymin=mean-std_err, ymax=mean+std_err), width=0.2) +
  labs(title="Random Forest CV Tuning — Hyperparameter Diagnostic",
       subtitle="Best mtry=3 at trees=400, interior to grid — range was appropriate",
       x="Number of Trees", y="CV AUC") +
  theme_minimal()

# XGBoost Diagnosis Plot — confirmed learn_rate range needed narrowing
gb_fit |>
  collect_metrics() |>
  filter(.metric == "roc_auc") |>
  mutate(tree_depth = as.factor(tree_depth)) |>
  ggplot(aes(x=learn_rate, y=mean, color=tree_depth)) +
  geom_point() + geom_line() +
  scale_x_log10() +
  geom_errorbar(aes(ymin=mean-std_err, ymax=mean+std_err), width=0.1) +
  labs(title="XGBoost CV Tuning — Hyperparameter Diagnostic",
       subtitle="Peak at learn_rate~0.006, interior to narrowed range — grid was appropriate",
       x="Learning Rate (log scale)", y="CV AUC") +
  theme_minimal()
```

**Performance metrics:** Area under the ROC curve (AUC) was the primary metric. Decision thresholds were selected using the Youden index (maximizing sensitivity + specificity - 1). Sensitivity and specificity at the optimal threshold are reported for each model.

# Results

## 5.1 Cohort Characteristics

The final analytic cohort comprised 4,179 adult sepsis ICU patients after applying all inclusion and exclusion criteria. Of these, 1,575 (37.7%) experienced prolonged ICU stays (≥14 days). The cohort had a median age of 65 years, was 57% male, and was predominantly White (61%). Medicare was the most common insurance type (52%). Hypertension (52%) and cardiovascular disease (38%) were the most prevalent comorbidities. Vasopressor use in the first 24 hours was recorded in 34% of patients.

```{r}
# final cohort summary using gtsummary
sepsis_imputed |>
  mutate(
    prolonged_stay = factor(prolonged_stay, labels = c("Normal (<14 days)", "Prolonged (≥14 days)")),
    gender         = factor(gender, levels = c("F", "M"), 
                labels = c("Female", "Male")),
    insurance      = factor(insurance),
    race_collapsed = case_when(
      str_detect(race, "WHITE")    ~ "WHITE",
      str_detect(race, "BLACK")    ~ "BLACK",
      str_detect(race, "HISPANIC") ~ "HISPANIC",
      str_detect(race, "ASIAN")    ~ "ASIAN",
      race %in% c("UNKNOWN", "UNABLE TO OBTAIN",
                  "PATIENT DECLINED TO ANSWER") ~ "UNKNOWN",
      TRUE ~ "OTHER"
    ) |> as.factor()
  ) |>
  select(anchor_age, gender, race_collapsed, insurance,
         hypertension, diabetes, ckd, cvd, chronic_liver,
         immunosuppression, vasopressor, gcs_total,
         los_days, prolonged_stay) |>
  tbl_summary(
    by = prolonged_stay,
    statistic = list(
      all_continuous()  ~ "{median} ({p25}-{p75}); {mean} ({sd})",
      all_categorical() ~ "{n} ({p}%)"
    ),
    digits = list(all_continuous() ~ 1),
    label = list(
      anchor_age        ~ "Age (years)",
      gender            ~ "Gender",
      race_collapsed    ~ "Race",
      insurance         ~ "Insurance",
      hypertension      ~ "Hypertension",
      diabetes          ~ "Diabetes",
      ckd               ~ "Chronic Kidney Disease (CKD)",
      cvd               ~ "Cardiovascular Disease (CVD)",
      chronic_liver     ~ "Chronic Liver Disease",
      immunosuppression ~ "Immunosuppression",
      vasopressor       ~ "Vasopressor Use",
      gcs_total         ~ "Glasgow Coma Scale (GCS) Total",
      los_days          ~ "Length of Stay (LOS, days)"
    )
  ) |>
add_p() |>
  add_overall() |>
  bold_labels() |>
  modify_footnote(
    all_stat_cols() ~ "Continuous: Median (Q1-Q3); Mean (SD). Categorical: n (%)"
  )

# Density Plot of continuous variables
sepsis_imputed |>
  pivot_longer(
    cols = c(gcs_total, creatinine, wbc,
             temperature, map, anchor_age,
             heart_rate, resp_rate, platelets),
    names_to  = "variable",
    values_to = "value"
  ) |>
  mutate(
    prolonged_stay = factor(prolonged_stay,
                            labels = c("Normal (<14 days)", "Prolonged (≥14 days)")),
    variable = recode(variable,
      gcs_total   = "Glasgow Coma Scale (GCS) Total",
      creatinine  = "Creatinine (mg/dL)",
      wbc         = "WBC (K/uL)",
      temperature = "Temperature (deg C)",
      map         = "MAP (mmHg)",
      anchor_age  = "Age (years)",
      heart_rate  = "Heart Rate (bpm)",
      resp_rate   = "Respiratory Rate (/min)",
      platelets   = "Platelets (K/uL)"
    )
  ) |>
  ggplot(aes(x = value, fill = prolonged_stay)) +
  geom_density(alpha = 0.5) +
  facet_wrap(~ variable, scales = "free", ncol = 3) +
  scale_fill_manual(values = c("aquamarine3", "salmon")) +
  labs(title = "Distribution of Key Continuous Predictors by Outcome",
       x = NULL, y = "Density", fill = "Outcome") +
  theme_minimal() +
  theme(legend.position = "bottom",
        strip.text = element_text(size = 9),
        panel.spacing = unit(1, "lines"))

# Comorbidity Prevalence by Outcome
sepsis_imputed |>
  mutate(prolonged_stay = factor(prolonged_stay,
                                  labels = c("Normal (<14 days)", 
                                             "Prolonged (≥14 days)"))) |>
  group_by(prolonged_stay) |>
  summarise(across(c(hypertension, diabetes, ckd, cvd,
                     chronic_liver, immunosuppression, vasopressor),
                   ~ round(mean(.x) * 100, 1))) |>
  pivot_longer(-prolonged_stay,
               names_to = "comorbidity",
               values_to = "pct") |>
  mutate(comorbidity = recode(comorbidity,
    hypertension      = "Hypertension",
    diabetes          = "Diabetes",
    ckd               = "Chronic Kidney Disease (CKD)",
    cvd               = "Cardiovascular Disease (CVD)",
    chronic_liver     = "Chronic Liver Disease",
    immunosuppression = "Immunosuppression",
    vasopressor       = "Vasopressor Use"
  )) |>
  ggplot(aes(x = reorder(comorbidity, pct), y = pct, fill = prolonged_stay)) +
  geom_col(position = "dodge", width = 0.6) +
  coord_flip() +
  scale_fill_manual(values = c("aquamarine3", "salmon")) +
  labs(title = "Comorbidity Prevalence by Outcome",
       x = NULL, y = "Prevalence (%)", fill = "Outcome") +
  theme_minimal() +
  theme(legend.position = "bottom")
```

## 5.2 Model Performance

All six models demonstrated comparable discriminative performance on the held-out test set (AUC range 0.677–0.683). XGBoost achieved the highest test AUC (0.683) following diagnostic-guided hyperparameter refinement. Ridge regression (0.681) and LASSO (0.680) performed comparably. Random Forest achieved an AUC of 0.677, modestly below the regularized regression models.

```{r}
#[INSERT TABLE: Model Performance Table (AUC, Threshold, Sensitivity, Specificity)] 
```

```{r}
# build roc_df from test set ROC objects
lab_lr    <- paste0("Logistic Regression\nAUC = ", round(auc_lr, 3))
lab_ridge <- paste0("Ridge\nAUC = ", round(auc_ridge, 3))
lab_lasso <- paste0("LASSO\nAUC = ", round(auc_lasso, 3))
lab_enet  <- paste0("Elastic Net\nAUC = ", round(auc_enet, 3))

roc_df <- bind_rows(
  data.frame(FPR = 1 - roc_lr$specificities,    TPR = roc_lr$sensitivities,    Model = lab_lr),
  data.frame(FPR = 1 - roc_ridge$specificities,  TPR = roc_ridge$sensitivities,  Model = lab_ridge),
  data.frame(FPR = 1 - roc_lasso$specificities,  TPR = roc_lasso$sensitivities,  Model = lab_lasso),
  data.frame(FPR = 1 - roc_enet$specificities,   TPR = roc_enet$sensitivities,   Model = lab_enet)
)

# combined all 4 into one graph
ggplot(roc_df, aes(x = FPR, y = TPR, color = Model)) +
  geom_line(lwd = 1) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "gray50") +
  scale_color_manual(values = setNames(
    c("yellow", "blue", "green", "red"),
    c(lab_lr, lab_ridge, lab_lasso, lab_enet)
  )) +
  labs(
    title = "ROC Curves by Model",
    x     = "False Positive Rate (1 - Specificity)",
    y     = "True Positive Rate (Sensitivity)"
  ) +
  theme_minimal() +
  theme(legend.position = "bottom")
```

The narrow range of AUC values across all models (0.006 difference between best and worst) suggests that the relationships between predictors and prolonged stay are predominantly linear, and that increasing model complexity did not yield meaningful performance gains in this dataset.

## 5.3 Feature Importance

LASSO variable selection retained 23 of 28 model matrix columns, dropping race_collapsedASIAN, race_collapsedUNKNOWN, diabetes (main effect), immunosuppression, and insurancePrivate. Notably, all three interaction terms were retained, with the CKD×Diabetes interaction kept while diabetes as a main effect was dropped — consistent with diabetic nephropathy as the clinically meaningful entity rather than diabetes alone. Chronic liver disease and cardiovascular disease had the largest positive LASSO coefficients, indicating the strongest association with prolonged stay. Insurance: Unknown had the largest negative coefficient, likely reflecting documentation patterns rather than true clinical differences.

```{r}
#nneeed my coefficient plot!!
#[INSERT FIGURE: LASSO Coefficient Plot] 
```

Random Forest and XGBoost variable importance plots both identified GCS total, age, and platelets as the top continuous predictors, consistent with LASSO findings.

```{r}
#[INSERT FIGURE: Random Forest Variable Importance Plot] 

#[INSERT FIGURE: XGBoost Variable Importance Plot]
```

## 8. Graphs & Tables

```{r}
# raw data: sepsis_with_labs --> outcome distribution
sepsis_with_labs |>
    count(prolonged_stay) |>
    mutate(label = if_else(prolonged_stay == 1, "Prolonged (>14 days)", "Normal (≤14 days)"),
           pct = round(n / sum(n) * 100, 1)) |>
    ggplot(aes(x = label, y = n, fill = label)) +
    geom_col(width = 0.5) +
    geom_text(aes(label = paste0(n, " (", pct, "%)")), 
              vjust = -0.5, size = 4) +
    scale_fill_manual(values = c("Prolonged (>14 days)" = "salmon",
                                 "Normal (≤14 days)"    = "aquamarine2")) +
    labs(title = "Pre-Imputation: Outcome Distribution: Prolonged Hospital Stay",
         x = NULL, y = "Number of Patients") +
    theme_minimal() +
    theme(legend.position = "none")

# sepsis_imputed: outcome distribution
sepsis_imputed |>
    count(prolonged_stay) |>
    mutate(label = if_else(prolonged_stay == 1, "Prolonged (>14 days)", "Normal (≤14 days)"),
           pct = round(n / sum(n) * 100, 1)) |>
    ggplot(aes(x = label, y = n, fill = label)) +
    geom_col(width = 0.5) +
    geom_text(aes(label = paste0(n, " (", pct, "%)")), 
              vjust = -0.5, size = 4) +
    scale_fill_manual(values = c("Prolonged (>14 days)" = "salmon",
                                 "Normal (≤14 days)"    = "aquamarine")) +
    labs(title = "Post-Imputation: Outcome Distribution: Prolonged Hospital Stay",
         x = NULL, y = "Number of Patients") +
    theme_minimal() +
    theme(legend.position = "none")
```

```{r}
#saving to rds file
app_data <- list(
  cohort      = sepsis_imputed,
  lasso_model = lasso_model
)

saveRDS(app_data, "/Users/charlottelee/Documents/biostat-203b-2026-winter/final_project/app_data.rds")
cat("Saved! File size:", 
    round(file.size("/Users/charlottelee/Documents/biostat-203b-2026-winter/final_project/app_data.rds")/1e6, 1), 
    "MB\n")
```
