---
title: "Machine Learning Prediction of Prolonged ICU Stay in Sepsis Patients Using MIMIC-IV"
subtitle: "Biostat 203B Final Project"
author: 
  - name: "Charlotte Lee and 206782165"
  - name: "Yasmine Tani and Brandon Brushwyler"
date: "March 20, 2026"
execute:
  cache: true
format:
  html:
    theme: cosmo
    embed-resources: true
    number-sections: false
    toc: true
    toc-depth: 4
    toc-location: left
    code-fold: false
---
# Abstract

250-300 words here

# Introduction
## Background
## Literature Review  
## Problem Statement
## Objectives

# Methods
## Data Source
Display machine information:

```{r}
sessionInfo()
```

Display my machine memory.

```{r}
memuse::Sys.meminfo()
```

Load database libraries and the tidyverse frontend:

```{r}
#| label: setup
#| cache: false
#| message: false
#| warning: false
library(bigrquery)
library(dbplyr)
library(DBI)
library(gt)
library(gtsummary)
library(tidyverse)
library(glmnet)
library(pROC)
library(corrplot)
library(tidymodels)
library(ranger)
library(vip)
library(xgboost)
library(styler)

```

## Connect to The Data

```{r}
#| cache: false  # overrides the YAML cache: true for this chunk only

bq_auth(path = "biostat-203b-2026-winter-92fefbfab477.json")
con_bq <- dbConnect(
  bigrquery::bigquery(),
  project = "biostat-203b-2025-winter",
  dataset = "mimiciv_3_1",
  billing = "biostat-203b-2025-winter"
)
```

# Data Cleaning + Pre-Processing

## 1. Filtering out Age (\>=18) & Primary Sepsis Diagnosis (upon admission) & ICU patients only

```{r}
# load raw tables
diagnoses_icd_tbl <- tbl(con_bq, "diagnoses_icd") |>
  arrange(subject_id, hadm_id)

d_icd_diagnoses_tbl <- tbl(con_bq, "d_icd_diagnoses")

patients_tbl <- tbl(con_bq, "patients") |>
  select(subject_id, anchor_age, gender)

# match sepsis icd codes from lookup
sepsis_codes_tbl <- d_icd_diagnoses_tbl |>
  filter(sql("LOWER(long_title) LIKE '%sepsis%'")) |>
  select(icd_code, icd_version, long_title)

sepsis_codes_tbl |> collect() |> print(n = 50)

# keep only primary diagnosis (seq_num == 1)
sepsis_cohort_tbl <- diagnoses_icd_tbl |>
  filter(seq_num == 1) |>
  inner_join(sepsis_codes_tbl, by = c("icd_code", "icd_version")) |>
  select(subject_id, hadm_id, seq_num, icd_code, icd_version, long_title)

# filter adults only
adults_tbl <- patients_tbl |>
  filter(anchor_age >= 18)

sepsis_adult_cohort_tbl <- sepsis_cohort_tbl |>
  inner_join(adults_tbl, by = "subject_id")

cohort_df <- sepsis_adult_cohort_tbl |> collect()

# restrict to ICU patients only upfront
icu_hadm_ids <- tbl(con_bq, "icustays") |>
  select(hadm_id) |>
  distinct() |>
  collect() |>
  pull(hadm_id)

cohort_df <- cohort_df |>
  filter(hadm_id %in% icu_hadm_ids)

# check:
cat("Total admissions: ", nrow(cohort_df), "\n")
cat("Unique patients:  ", n_distinct(cohort_df$subject_id), "\n")
print(summary(cohort_df$anchor_age))

```

## 2. Outcome Variable + 3. Adding Predictors

```{r}
# pulls LOS + race + insurance in one query instead of two
admissions_tbl <- tbl(con_bq, "admissions") |>
  select(subject_id, hadm_id, admittime, dischtime, race, insurance) |>
  mutate(los_days = sql("TIMESTAMP_DIFF(dischtime, admittime, DAY)")) |>
  collect()

# join onto cohort and create outcome variable (prolonged stay >= 14 days)
sepsis_with_outcome <- cohort_df |>
  inner_join(admissions_tbl, by = c("subject_id", "hadm_id")) |>
  filter(!is.na(los_days), los_days > 0) |>
  mutate(prolonged_stay = if_else(los_days >= 14, 1L, 0L))

# check:
cat("Total admissions:       ", nrow(sepsis_with_outcome), "\n")
cat("Unique patients:        ", n_distinct(sepsis_with_outcome$subject_id), "\n")
cat("Prolonged stay (>=14d): ", sum(sepsis_with_outcome$prolonged_stay), "\n")
cat("% prolonged:            ", round(mean(sepsis_with_outcome$prolonged_stay)*100, 1), "%\n")
cat("Median LOS:             ", median(sepsis_with_outcome$los_days), "days\n")

# insurance check
cat("\n--- Insurance Distribution ---\n")
table(sepsis_with_outcome$insurance) |> print()
cat("% missing insurance:    ", round(mean(is.na(sepsis_with_outcome$insurance))*100, 1), "%\n")
sepsis_with_demo <- sepsis_with_outcome

```

### 3.2 Comorbidities

```{r}
cohort_ids <- sepsis_with_demo$hadm_id

# pull secondary diagnoses — both ICD-9 AND ICD-10 to capture all admissions
# collect AFTER all BQ-side filters
secondary_dx <- tbl(con_bq, "diagnoses_icd") |>
  filter(
    seq_num > 1,
    hadm_id %in% !!cohort_ids
  ) |>
  select(subject_id, hadm_id, icd_code, icd_version) |>
  collect()

comorbidity_flags <- secondary_dx |>
  mutate(
    # hypertension — ICD-10: I10 | ICD-9: 401-405
    hypertension = if_else(
      (icd_version == 10 & startsWith(icd_code, "I10")) |
      (icd_version == 9  & (startsWith(icd_code, "401") | startsWith(icd_code, "402") |
                             startsWith(icd_code, "403") | startsWith(icd_code, "404") |
                             startsWith(icd_code, "405"))),
      1L, 0L),

    # diabetes — ICD-10: E10/E11 | ICD-9: 250
    diabetes = if_else(
      (icd_version == 10 & (startsWith(icd_code, "E10") | startsWith(icd_code, "E11"))) |
      (icd_version == 9  &  startsWith(icd_code, "250")),
      1L, 0L),

    # chronic kidney disease — ICD-10: N18 | ICD-9: 585-586
    ckd = if_else(
      (icd_version == 10 & startsWith(icd_code, "N18")) |
      (icd_version == 9  & (startsWith(icd_code, "585") | startsWith(icd_code, "586"))),
      1L, 0L),

    # cardiovascular disease — ICD-10: I20/I21/I25/I48/I50/I70/I71
    #                           ICD-9: 410-414/427/428/440/441
    cvd = if_else(
      (icd_version == 10 & (
        startsWith(icd_code, "I20") | startsWith(icd_code, "I21") |
        startsWith(icd_code, "I25") | startsWith(icd_code, "I48") |
        startsWith(icd_code, "I50") | startsWith(icd_code, "I70") |
        startsWith(icd_code, "I71"))) |
      (icd_version == 9  & (
        startsWith(icd_code, "410") | startsWith(icd_code, "411") |
        startsWith(icd_code, "412") | startsWith(icd_code, "413") |
        startsWith(icd_code, "414") | startsWith(icd_code, "427") |
        startsWith(icd_code, "428") | startsWith(icd_code, "440") |
        startsWith(icd_code, "441"))),
      1L, 0L),

    # chronic liver disease — ICD-10: K70/K73/K74 | ICD-9: 571
    chronic_liver = if_else(
      (icd_version == 10 & (
        startsWith(icd_code, "K70") | startsWith(icd_code, "K73") |
        startsWith(icd_code, "K74"))) |
      (icd_version == 9  &  startsWith(icd_code, "571")),
      1L, 0L),

    # immunosuppression — ICD-10: D80/D81/D83/Z79/Z94/C81-C83/C91/C92
    #                     ICD-9: 279/V42/V45/200-202/204/205
    immunosuppression = if_else(
      (icd_version == 10 & (
        startsWith(icd_code, "D80") | startsWith(icd_code, "D81") |
        startsWith(icd_code, "D83") | startsWith(icd_code, "Z79") |
        startsWith(icd_code, "Z94") | startsWith(icd_code, "C81") |
        startsWith(icd_code, "C82") | startsWith(icd_code, "C83") |
        startsWith(icd_code, "C91") | startsWith(icd_code, "C92"))) |
      (icd_version == 9  & (
        startsWith(icd_code, "279") | startsWith(icd_code, "V42") |
        startsWith(icd_code, "V45") | startsWith(icd_code, "200") |
        startsWith(icd_code, "201") | startsWith(icd_code, "202") |
        startsWith(icd_code, "204") | startsWith(icd_code, "205"))),
      1L, 0L)
  ) |>
  group_by(subject_id, hadm_id) |>
  summarise(across(hypertension:immunosuppression, max), .groups = "drop") |>
  mutate(across(hypertension:immunosuppression, ~replace_na(.x, 0L)))

sepsis_with_comorbidities <- sepsis_with_demo |>
  left_join(comorbidity_flags, by = c("subject_id", "hadm_id")) |>
  mutate(across(hypertension:immunosuppression, ~replace_na(.x, 0L)))

# check 
cat("Rows after joining comorbidities: ", nrow(sepsis_with_comorbidities), "\n")
comorbidity_flags |>
  summarise(across(hypertension:immunosuppression, ~round(mean(.x)*100,1))) |>
  print(width=Inf)
dim(sepsis_with_comorbidities)
```

### 3.3 ICU admission Times

```{r}
### 3.3 ICU Admission Times (for 24h vital/lab windows)

cohort_ids <- sepsis_with_comorbidities$hadm_id

# get earliest ICU admission time per stay — used to define 24h window below
icu_admissions_bq <- tbl(con_bq, "icustays") |>
  filter(hadm_id %in% !!cohort_ids) |>
  group_by(subject_id, hadm_id) |>
  summarise(intime = min(intime, na.rm = TRUE), .groups = "drop")

icustays_raw <- icu_admissions_bq |> collect()
icu_admissions <- icustays_raw

cat("ICU stays pulled:", nrow(icustays_raw), "\n")
```

### 3.4 Vital Items

which map is being used here (bc there's 2)

```{r}
# pull vitals from chartevents within first 24h of ICU admission
# itemid 223761 is fahrenheit so converting to celsius
vitals_raw <- tbl(con_bq, "chartevents") |>
  filter(
    itemid %in% c(220045, 220179, 220050, 220180, 220051,
                  220052, 220181, 220210, 223762, 223761),
    hadm_id %in% !!cohort_ids,
    !is.na(valuenum)
  ) |>
  select(subject_id, hadm_id, itemid, charttime, valuenum) |>
  inner_join(icu_admissions_bq, by = c("subject_id", "hadm_id")) |>
  filter(
    charttime >= intime,
    charttime <= sql("TIMESTAMP_ADD(intime, INTERVAL 24 HOUR)")
  ) |>
  mutate(
    valuenum = if_else(itemid == 223761, (valuenum - 32) * 5/9, valuenum)
  ) |>
  collect()   # now collects only 24h window rows, not the whole table

# map itemids to vital names (some vitals have multiple itemids)
vital_items <- tibble::tibble(
  itemid     = c(220045, 220179, 220050, 220180, 220051,
                 220052, 220181, 220210, 223762, 223761),
  vital_name = c("heart_rate", "sbp",  "sbp",  "dbp",  "dbp",
                 "map",  "map",  "resp_rate", "temperature", "temperature")
)
  #map combined then min of both found which is clincially standard approach

# summarise each vital per admission (worst values)
vitals_summary <- vitals_raw |>
  left_join(vital_items, by = "itemid") |>
  group_by(subject_id, hadm_id, vital_name) |>
  summarise(
    value = case_when(
      # MAX — tachycardia indicates cardiac stress/infection severity
      first(vital_name) == "heart_rate"  ~ max(valuenum),
      
      # MIN — hypotension indicates shock/hemodynamic instability
      first(vital_name) == "sbp"         ~ min(valuenum),
      first(vital_name) == "dbp"         ~ min(valuenum),
      first(vital_name) == "map"         ~ min(valuenum),
      
      # MAX — tachypnea indicates respiratory distress/metabolic acidosis
      first(vital_name) == "resp_rate"   ~ max(valuenum),
      
      # MAX — peak fever indicates infection severity
      # (hypothermia in sepsis also dangerous but less common)
      first(vital_name) == "temperature" ~ max(valuenum),
    ),
    .groups = "drop"
  ) |>
  pivot_wider(names_from = vital_name, values_from = value)
# check
cat("Rows after joining vitals:", nrow(vitals_summary), "\n")

# may take up to a minute to run
```

### 3.5 Lab Events

```{r}
#Confirm Lab Items: 
# verify all lab itemids are correct
#tbl(con_bq, "d_labitems") |>
 # filter(itemid %in% c(50912, 50885, 51301, 51265, 50813)) |>
#  select(itemid, label) |>
#  arrange(itemid) |>
#  collect() |>
#  print()

# pull labs from labevents within first 24h of ICU admission
labs_raw <- tbl(con_bq, "labevents") |>
  filter(
    itemid %in% c(50912, 50885, 51301, 51265, 50813),
    hadm_id %in% !!cohort_ids,
    !is.na(valuenum)
  ) |>
  select(subject_id, hadm_id, itemid, charttime, valuenum) |>
  inner_join(icu_admissions_bq, by = c("subject_id", "hadm_id")) |>
  filter(
    charttime >= intime,
    charttime <= sql("TIMESTAMP_ADD(intime, INTERVAL 24 HOUR)")
  ) |>
  collect()   # collect AFTER all filters

# map itemids to lab names
lab_items <- tibble::tibble(
  itemid   = c(50912,        50885,       51301,  51265,       50813),
  lab_name = c("creatinine", "bilirubin", "wbc",  "platelets", "lactate")
)

# worst values per admission
labs_summary <- labs_raw |>
  left_join(lab_items, by = "itemid") |>
  group_by(subject_id, hadm_id, lab_name) |>
  summarise(
    value = case_when(
      # MAX — peak creatinine indicates worst kidney function
      # (higher creatinine = more kidney damage)
      first(lab_name) == "creatinine" ~ max(valuenum),
      
      # MAX — peak bilirubin indicates worst liver function
      # (higher bilirubin = more liver damage/dysfunction)
      first(lab_name) == "bilirubin"  ~ max(valuenum),
      
      # MAX — peak WBC indicates worst infection/inflammatory response
      # (very high WBC = severe infection)
      first(lab_name) == "wbc"        ~ max(valuenum),
      
      # MIN — lowest platelets is most dangerous
      # (low platelets = coagulopathy/DIC risk, common in severe sepsis)
      first(lab_name) == "platelets"  ~ min(valuenum),
      
      # MAX — peak lactate indicates worst tissue perfusion
      # (high lactate = cells not getting enough oxygen = septic shock)
      first(lab_name) == "lactate"    ~ max(valuenum),
    ),
    .groups = "drop"
  ) |>
  pivot_wider(names_from = lab_name, values_from = value)

# check
cat("Rows after joining labs:", nrow(labs_summary), "\n")
```

### 3.6 GCS (first 24 hr minimum)

```{r}
### 3.6 GCS (first 24h minimum — lowest score = worst neurological state)
gcs_raw <- tbl(con_bq, "chartevents") |>
  filter(
    itemid %in% c(220739, 223900, 223901),
    hadm_id %in% !!cohort_ids,
    !is.na(valuenum)
  ) |>
  select(subject_id, hadm_id, itemid, charttime, valuenum) |>
  inner_join(icu_admissions_bq, by = c("subject_id", "hadm_id")) |>
  filter(
    charttime >= intime,
    charttime <= sql("TIMESTAMP_ADD(intime, INTERVAL 24 HOUR)")
  ) |>
  collect() |>
  group_by(subject_id, hadm_id, itemid) |>
  summarise(value = min(valuenum, na.rm = TRUE), .groups = "drop") |>
  pivot_wider(names_from = itemid, values_from = value) |>
  rename(gcs_eye    = `220739`,
         gcs_verbal = `223900`,
         gcs_motor  = `223901`) |>
  mutate(gcs_total = gcs_eye + gcs_verbal + gcs_motor) |>
  select(subject_id, hadm_id, gcs_total)

cat("Patients with GCS data:", nrow(gcs_raw), "\n")
summary(gcs_raw$gcs_total)
```

### 3.7 Vasopressors (first 24h binary flags)

```{r}
# vasopressor itemids — norepinephrine, epinephrine, dopamine, vasopressin, phenylephrine
vasopressor_ids <- c(221906, 221289, 221662, 222315, 221749)

vasopressors_raw <- tbl(con_bq, "inputevents") |>
  filter(
    itemid %in% vasopressor_ids,
    hadm_id %in% !!cohort_ids
  ) |>
  select(subject_id, hadm_id, itemid, starttime) |>
  inner_join(icu_admissions_bq, by = c("subject_id", "hadm_id")) |>
  filter(
    starttime >= intime,
    starttime <= sql("TIMESTAMP_ADD(intime, INTERVAL 24 HOUR)")
  ) |>
  collect()

# binary flag — 1 if patient received vasopressors in first 24h, 0 otherwise
vasopressor_flags <- vasopressors_raw |>
  group_by(subject_id, hadm_id) |>
  summarise(vasopressor = 1L, .groups = "drop")

# check
cat("Patients with vasopressors in first 24h:", nrow(vasopressor_flags), "\n")
```

### 3.8 Safety Check + Joining all Predictors

```{r}
### 3.8 Safety Check + Joining all Predictors

# ICU filter already applied in section 1 — this is just a safety check
sepsis_icu <- sepsis_with_comorbidities

sepsis_with_vitals <- sepsis_icu |>
  left_join(vitals_summary, by = c("subject_id", "hadm_id"))

sepsis_with_labs <- sepsis_with_vitals |>
  left_join(labs_summary,      by = c("subject_id", "hadm_id")) |>
  left_join(vasopressor_flags, by = c("subject_id", "hadm_id")) |>
  left_join(gcs_raw,           by = c("subject_id", "hadm_id")) |>  # added GCS
  mutate(vasopressor = replace_na(vasopressor, 0L)) |>
  select(-sbp, -dbp)   # dropped sbp monitor 1, sdp monitor 2, dbp (both monitors), sbpmulticollinear variables

cat("Final rows:", nrow(sepsis_with_labs), "\n")
cat("Vasopressor use (%):", round(mean(sepsis_with_labs$vasopressor)*100, 1), "\n")
cat("GCS missing (%):", round(mean(is.na(sepsis_with_labs$gcs_total))*100, 1), "\n")
```

## 4. Raw Data

```{r}
#raw data view
summary(sepsis_with_labs)

# check
cat("raw final dataset:", nrow(sepsis_icu), "\n")
cat("Final rows:", nrow(sepsis_with_labs), "\n")

# check % missingness across all columns
sepsis_with_labs %>%
  summarise(across(everything(), ~ round(mean(is.na(.x)) * 100, 1))) %>%
  print(width = Inf)
```

## 5. Outlier Cleaning

```{r}
# Outlier cleaning note:
# bounds set at physiologically IMPOSSIBLE limits, not clinical abnormality limits
# goal is to remove data entry errors while preserving true disease severity
# e.g. lactate of 28 mmol/L is a real critically ill patient — kept
#      lactate of 1,276,103 is clearly a data entry error — removed
# SBP/MAP/DBP lower bound set to >0 (not >=0) since zero BP = death/error
# resp_rate max 73, wbc max 267, lactate max 29.6 retained as physiologically 
# possible extreme values in severely ill sepsis patients
# temperature max 42.5C retained as real severe fever


# remove physiologically IMPOSSIBLE values only — not clinical abnormalities
# goal: remove data entry errors, NOT filter out severely ill patients
# a lactate of 28 is real and kept; a lactate of 1,276,103 is a data error
sepsis_clean <- sepsis_with_labs |>
  mutate(
    # heart rate: no human survives >300, <20 is incompatible with life
    heart_rate  = if_else(heart_rate > 300 | heart_rate < 20,
                          NA_real_, heart_rate),

    # map: zero or negative impossible, >200 is data error
    map         = if_else(map <= 0 | map > 200, NA_real_, map),

    # resp rate: <5 impossible, >80 is data error
    resp_rate   = if_else(resp_rate < 5 | resp_rate > 80,
                          NA_real_, resp_rate),

    # temperature: <25 incompatible with life, >45 lethal/data error
    temperature = if_else(temperature < 25 | temperature > 45,
                          NA_real_, temperature),

    # lactate: >30 is beyond any recorded human value
    lactate     = if_else(lactate < 0 | lactate > 30, NA_real_, lactate),

    # wbc: >500 is beyond any recorded human value
    wbc         = if_else(wbc > 500, NA_real_, wbc),

    # platelets: >3000 is beyond any recorded human value
    platelets   = if_else(platelets > 3000, NA_real_, platelets),

    # bilirubin: >150 is beyond any recorded human value
    bilirubin   = if_else(bilirubin > 150, NA_real_, bilirubin),

    # creatinine: >50 is beyond any recorded human value
    creatinine  = if_else(creatinine > 50, NA_real_, creatinine),

    # GCS: scale is 3-15 by definition, outside = data error
    gcs_total   = if_else(gcs_total < 3 | gcs_total > 15,
                          NA_real_, gcs_total)
  )

# insurance doesn't need outlier cleaning — categorical with fixed levels

# check missingness after cleaning
sepsis_clean |>
  summarise(across(everything(), ~ round(mean(is.na(.x)) * 100, 1))) |>
  print(width = Inf)

# verify crazy values are gone
summary(sepsis_clean[, c("resp_rate", "lactate", "temperature",
                          "map", "wbc", "creatinine", "gcs_total")])

```

## 6. Imputation

```{r}
# step 1: keep first admission per patient
sepsis_clean <- sepsis_clean |>
  arrange(subject_id, admittime) |>
  group_by(subject_id) |>
  slice(1) |>
  ungroup()

cat("After first admission filter:", nrow(sepsis_clean), "\n")
cat("Duplicate subject_ids:", sum(duplicated(sepsis_clean$subject_id)), "\n")

# step 2: remove patients who died within 48h of ICU admission
# these patients cannot have prolonged stay by definition
# and create competing risk bias — sickest patients die quickly,
# making high lactate/vasopressor falsely appear protective
early_deaths <- tbl(con_bq, "admissions") |>
  select(subject_id, hadm_id, hospital_expire_flag, deathtime) |>
  collect()

sepsis_clean <- sepsis_clean |>
  left_join(early_deaths, by = c("subject_id", "hadm_id")) |>
  left_join(icustays_raw |> select(subject_id, hadm_id, intime),
            by = c("subject_id", "hadm_id")) |>
  filter(
    hospital_expire_flag == 0 |
    is.na(deathtime) |
    as.numeric(difftime(deathtime, intime, units = "hours")) > 48
  ) |>
  select(-hospital_expire_flag, -deathtime, -intime)

cat("After removing early deaths:", nrow(sepsis_clean), "\n")

# step 3: fix Inf/-Inf then median impute remaining NAs
# Inf arises when max/min was called on a patient with ALL NA values in 24h window
# median chosen over mean due to right-skewed distributions in clinical variables
# gcs_total: low score = worst neurological state, median impute missing values
sepsis_imputed <- sepsis_clean |>
  mutate(across(
    .cols = c(heart_rate, map, resp_rate, temperature,
              creatinine, bilirubin, wbc, platelets, lactate, gcs_total),
    ~ if_else(is.infinite(.x), NA_real_, .x)
  )) |>
  mutate(across(
    .cols = c(heart_rate, map, resp_rate, temperature,
              creatinine, bilirubin, wbc, platelets, lactate, gcs_total),
    ~ coalesce(.x, median(.x, na.rm = TRUE))
  ))

# confirm no remaining missingness
sepsis_imputed |>
  summarise(across(c(heart_rate, map, resp_rate, temperature,
                     creatinine, bilirubin, wbc, platelets, 
                     lactate, gcs_total),
                   ~ round(mean(is.na(.x)) * 100, 1))) |>
  print()

cat("Final patients:", nrow(sepsis_imputed), "\n")
cat("Duplicate subject_ids:", sum(duplicated(sepsis_imputed$subject_id)), "\n")

```

## 7. Final Cohort Table

```{r}
# final cohort summary using gtsummary
sepsis_imputed |>
  mutate(
    prolonged_stay = factor(prolonged_stay, labels = c("Normal (<14 days)", "Prolonged (≥14 days)")),
    gender         = factor(gender),
    insurance      = factor(insurance),
    race_collapsed = case_when(
      str_detect(race, "WHITE")    ~ "WHITE",
      str_detect(race, "BLACK")    ~ "BLACK",
      str_detect(race, "HISPANIC") ~ "HISPANIC",
      str_detect(race, "ASIAN")    ~ "ASIAN",
      race %in% c("UNKNOWN", "UNABLE TO OBTAIN",
                  "PATIENT DECLINED TO ANSWER") ~ "UNKNOWN",
      TRUE ~ "OTHER"
    ) |> as.factor()
  ) |>
  select(anchor_age, gender, race_collapsed, insurance,
         hypertension, diabetes, ckd, cvd, chronic_liver,
         immunosuppression, vasopressor, gcs_total,
         los_days, prolonged_stay) |>
  tbl_summary(
    by = prolonged_stay,
    label = list(
      anchor_age        ~ "Age (years)",
      gender            ~ "Gender",
      race_collapsed    ~ "Race",
      insurance         ~ "Insurance",
      hypertension      ~ "Hypertension",
      diabetes          ~ "Diabetes",
      ckd               ~ "CKD",
      cvd               ~ "CVD",
      chronic_liver     ~ "Chronic Liver Disease",
      immunosuppression ~ "Immunosuppression",
      vasopressor       ~ "Vasopressor Use",
      gcs_total         ~ "GCS Total (first 24h minimum)",
      los_days          ~ "Length of Stay (days)"
    )
  ) |>
  add_p() |>
  add_overall() |>
  bold_labels()
```

## 8. Graphs & Tables

```{r}
# raw data: sepsis_with_labs --> outcome distribution
sepsis_with_labs |>
    count(prolonged_stay) |>
    mutate(label = if_else(prolonged_stay == 1, "Prolonged (>14 days)", "Normal (≤14 days)"),
           pct = round(n / sum(n) * 100, 1)) |>
    ggplot(aes(x = label, y = n, fill = label)) +
    geom_col(width = 0.5) +
    geom_text(aes(label = paste0(n, " (", pct, "%)")), 
              vjust = -0.5, size = 4) +
    scale_fill_manual(values = c("Prolonged (>14 days)" = "salmon",
                                 "Normal (≤14 days)"    = "aquamarine2")) +
    labs(title = "Pre-Imputation: Outcome Distribution: Prolonged Hospital Stay",
         x = NULL, y = "Number of Patients") +
    theme_minimal() +
    theme(legend.position = "none")

# sepsis_imputed: outcome distribution
sepsis_imputed |>
    count(prolonged_stay) |>
    mutate(label = if_else(prolonged_stay == 1, "Prolonged (>14 days)", "Normal (≤14 days)"),
           pct = round(n / sum(n) * 100, 1)) |>
    ggplot(aes(x = label, y = n, fill = label)) +
    geom_col(width = 0.5) +
    geom_text(aes(label = paste0(n, " (", pct, "%)")), 
              vjust = -0.5, size = 4) +
    scale_fill_manual(values = c("Prolonged (>14 days)" = "salmon",
                                 "Normal (≤14 days)"    = "aquamarine")) +
    labs(title = "Post-Imputation: Outcome Distribution: Prolonged Hospital Stay",
         x = NULL, y = "Number of Patients") +
    theme_minimal() +
    theme(legend.position = "none")

# sepsis_imputed: age distribution
ggplot(sepsis_imputed, aes(x = anchor_age)) +
    geom_histogram(binwidth = 5, fill = "lightblue", color = "white") +
    geom_vline(xintercept = mean(sepsis_imputed$anchor_age), 
               linetype = "dashed", color = "red", linewidth = 0.8) +
    annotate("text", x = mean(sepsis_imputed$anchor_age) + 2, y = 300,
             label = paste0("Mean: ", round(mean(sepsis_imputed$anchor_age), 1)),
             color = "red", hjust = 0) +
    labs(title = "Age Distribution of ICU Sepsis Cohort",
         x = "Age (years)", y = "Count") +
    theme_minimal()

```

# Model 1:

## Logistic Regression: Lasso

```{r}
# race included as categorical predictor — dummy coded automatically by model.matrix
# gender recoded as binary (1 = M, 0 = F)
# vasopressor: binary flag — received any vasopressor in first 24h of ICU admission
# model matrix
#insurance: captures social determinants of health — meaningful variation in prolonged
#            stay across categories (Other 46.1% vs Medicare 34.1%)
model_vars <- c("anchor_age", "gender", "race_collapsed",
                "heart_rate", "map", "resp_rate",
                "temperature", "creatinine",
                "wbc", "platelets", "gcs_total",
                "hypertension", "diabetes", "ckd", "cvd",
                "chronic_liver", "immunosuppression",
                "vasopressor", "insurance")

model_df <- sepsis_imputed |>
  mutate(
    gender = if_else(gender == "M", 1L, 0L),
    race_collapsed = case_when(
      str_detect(race, "WHITE")    ~ "WHITE",
      str_detect(race, "BLACK")    ~ "BLACK",
      str_detect(race, "HISPANIC") ~ "HISPANIC",
      str_detect(race, "ASIAN")    ~ "ASIAN",
      race %in% c("UNKNOWN", "UNABLE TO OBTAIN",
                  "PATIENT DECLINED TO ANSWER") ~ "UNKNOWN",
      TRUE ~ "OTHER"
    ) |> as.factor(),
    insurance = if_else(is.na(insurance), "Unknown",
                        as.character(insurance)) |> as.factor()
  ) |>
  select(all_of(model_vars), prolonged_stay) |>
  mutate(across(where(is.numeric), ~if_else(is.infinite(.x), NA_real_, .x))) |>
  drop_na()

cat("Rows in model_df:", nrow(model_df), "\n")
cat("Total rows in sepsis_imputed:", nrow(sepsis_imputed), "\n")
cat("Rows dropped by drop_na():", nrow(sepsis_imputed) - nrow(model_df), "\n")
cat("Insurance distribution:\n")
table(model_df$insurance) |> print()
cat("Outcome distribution:\n")
prop.table(table(model_df$prolonged_stay)) |> round(3) |> print()

X <- model.matrix(~ . - prolonged_stay - 1, data = model_df)
y <- model_df$prolonged_stay

cat("Model matrix dimensions:", dim(X), "\n")
cat("Number of predictors:", ncol(X), "\n")

best_threshold <- function(roc_obj) {
  idx <- which.max(roc_obj$sensitivities + roc_obj$specificities - 1)
  roc_obj$thresholds[idx]
}

calc_metrics <- function(pred_prob, actual, threshold) {
  pred_class <- if_else(pred_prob >= threshold, 1L, 0L)
  sens <- mean(pred_class[actual == 1] == 1) * 100
  spec <- mean(pred_class[actual == 0] == 0) * 100
  c(sens, spec)
}

set.seed(42)
train_idx <- sample(1:nrow(X), 0.8 * nrow(X))
X_train <- X[train_idx, ]
X_test  <- X[-train_idx, ]
y_train <- y[train_idx]
y_test  <- y[-train_idx]

# add interaction terms to X:
X_interact <- model.matrix(
  ~ . - prolonged_stay - 1 + 
    chronic_liver:cvd +      # liver disease + cardiovascular disease
    ckd:diabetes +           # diabetic nephropathy
    gcs_total:vasopressor,   # neurological impairment + hemodynamic instability
  data = model_df
)

# use X_interact instead of X for all models
X_train_i <- X_interact[train_idx, ]
X_test_i  <- X_interact[-train_idx, ]
```

```{r}
# LASSO logistic regression: same as logistic regression but penalizes 
# small coefficients to zero — automatic variable selection
# Ridge: shrinks but keeps all predictors
# Elastic Net: mix of LASSO and Ridge
# LASSO chosen for variable selection despite similar AUC across all models

# plain logistic regression
set.seed(42)
lr_model <- glm(y_train ~ ., data = as.data.frame(X_train_i), family = "binomial")
lr_pred  <- predict(lr_model, newdata = as.data.frame(X_test_i), type = "response")

# LASSO
set.seed(42)
cv_lasso    <- cv.glmnet(X_train_i, y_train, family = "binomial", alpha = 1, nfolds = 10)
lasso_model <- glmnet(X_train_i, y_train, family = "binomial", alpha = 1, lambda = cv_lasso$lambda.min)
lasso_pred  <- as.numeric(predict(lasso_model, newx = X_test_i, type = "response"))

# ridge
set.seed(42)
cv_ridge    <- cv.glmnet(X_train_i, y_train, family = "binomial", alpha = 0, nfolds = 10)
ridge_model <- glmnet(X_train_i, y_train, family = "binomial", alpha = 0, lambda = cv_ridge$lambda.min)
ridge_pred  <- as.numeric(predict(ridge_model, newx = X_test_i, type = "response"))

# elastic net
set.seed(42)
cv_enet    <- cv.glmnet(X_train_i, y_train, family = "binomial", alpha = 0.5, nfolds = 10)
enet_model <- glmnet(X_train_i, y_train, family = "binomial", alpha = 0.5, lambda = cv_enet$lambda.min)
enet_pred  <- as.numeric(predict(enet_model, newx = X_test_i, type = "response"))

# compute ROC objects once
roc_lr    <- suppressMessages(roc(y_test, lr_pred))
roc_lasso <- suppressMessages(roc(y_test, lasso_pred))
roc_ridge <- suppressMessages(roc(y_test, ridge_pred))
roc_enet  <- suppressMessages(roc(y_test, enet_pred))

auc_lr    <- auc(roc_lr)
auc_lasso <- auc(roc_lasso)
auc_ridge <- auc(roc_ridge)
auc_enet  <- auc(roc_enet)

thresh_lr    <- best_threshold(roc_lr)
thresh_lasso <- best_threshold(roc_lasso)
thresh_ridge <- best_threshold(roc_ridge)
thresh_enet  <- best_threshold(roc_enet)

m_lr    <- calc_metrics(lr_pred,    y_test, thresh_lr)
m_lasso <- calc_metrics(lasso_pred, y_test, thresh_lasso)
m_ridge <- calc_metrics(ridge_pred, y_test, thresh_ridge)
m_enet  <- calc_metrics(enet_pred,  y_test, thresh_enet)

# clean results table
results_df <- tibble(
  Model       = c("Logistic Regression", "LASSO", "Ridge", "Elastic Net"),
  AUC         = round(c(auc_lr, auc_lasso, auc_ridge, auc_enet), 3),
  Threshold   = round(c(thresh_lr, thresh_lasso, thresh_ridge, thresh_enet), 3),
  Sensitivity = round(c(m_lr[1], m_lasso[1], m_ridge[1], m_enet[1]), 1),
  Specificity = round(c(m_lr[2], m_lasso[2], m_ridge[2], m_enet[2]), 1)
)
results_df |> gt()

# LASSO selected predictors
lasso_coefs <- coef(lasso_model) |>
  as.matrix() |>
  as.data.frame() |>
  rownames_to_column("Variable") |>
  filter(Variable != "(Intercept)") |>
  mutate(dropped = if_else(s0 == 0, "DROPPED", "kept"))

cat("Dropped by LASSO:\n")
lasso_coefs |> filter(dropped == "DROPPED") |> pull(Variable) |> print()

cat("\nKept by LASSO:\n") 
lasso_coefs |> filter(dropped == "kept")|> arrange(desc(abs(s0))) |> pull(Variable) |> print()
```

## Graphing ROC Curves

```{r}
# build roc_df from test set ROC objects
lab_lr    <- paste0("Logistic Regression\nAUC = ", round(auc_lr, 3))
lab_ridge <- paste0("Ridge\nAUC = ", round(auc_ridge, 3))
lab_lasso <- paste0("LASSO\nAUC = ", round(auc_lasso, 3))
lab_enet  <- paste0("Elastic Net\nAUC = ", round(auc_enet, 3))

roc_df <- bind_rows(
  data.frame(FPR = 1 - roc_lr$specificities,    TPR = roc_lr$sensitivities,    Model = lab_lr),
  data.frame(FPR = 1 - roc_ridge$specificities,  TPR = roc_ridge$sensitivities,  Model = lab_ridge),
  data.frame(FPR = 1 - roc_lasso$specificities,  TPR = roc_lasso$sensitivities,  Model = lab_lasso),
  data.frame(FPR = 1 - roc_enet$specificities,   TPR = roc_enet$sensitivities,   Model = lab_enet)
)

# combined all 4 into one graph
ggplot(roc_df, aes(x = FPR, y = TPR, color = Model)) +
  geom_line(lwd = 1) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "gray50") +
  scale_color_manual(values = setNames(
    c("yellow", "blue", "green", "red"),
    c(lab_lr, lab_ridge, lab_lasso, lab_enet)
  )) +
  labs(
    title = "ROC Curves by Model",
    x     = "False Positive Rate (1 - Specificity)",
    y     = "True Positive Rate (Sensitivity)"
  ) +
  theme_minimal() +
  theme(legend.position = "bottom")

```

# Random Forest for Classification Model

```{r}
# ── Random Forest ─────────────────────────────────────────────────────────────
set.seed(42)
sepsis_split <- initial_split(
  model_df |> mutate(prolonged_stay = as.factor(prolonged_stay)),
  prop = 0.8,
  strata = "prolonged_stay"
)
sepsis_other <- training(sepsis_split)
sepsis_test  <- testing(sepsis_split)

rf_recipe <- recipe(prolonged_stay ~ ., data = sepsis_other) |>
  step_dummy(all_nominal_predictors()) |>
  step_zv(all_numeric_predictors())

rf_mod <- rand_forest(
  mode  = "classification",
  mtry  = tune(),
  trees = tune()
) |>
  set_engine("ranger", importance = "impurity")

rf_wf <- workflow() |>
  add_recipe(rf_recipe) |>
  add_model(rf_mod)

# updated: trees expanded to 200-500 (plot showed upward trend at 300)
# mtry 1-5 unchanged — best mtry=3 was interior to range
param_grid_rf <- grid_regular(
  trees(range = c(200L, 500L)),
  mtry(range  = c(1L, 5L)),
  levels = c(4, 5)
)

set.seed(42)
folds <- vfold_cv(sepsis_other, v = 5, strata = "prolonged_stay")

rf_fit <- rf_wf |>
  tune_grid(
    resamples = folds,
    grid      = param_grid_rf,
    metrics   = metric_set(roc_auc, accuracy)
  )

# diagnosis plot
rf_fit |>
  collect_metrics() |>
  filter(.metric == "roc_auc") |>
  mutate(mtry = as.factor(mtry)) |>
  ggplot(aes(x = trees, y = mean, color = mtry, group = mtry)) +
  geom_line() + geom_point() +
  geom_errorbar(aes(ymin=mean-std_err, ymax=mean+std_err), width=10) +
  labs(x = "Number of Trees", y = "CV AUC",
       title = "Random Forest CV Tuning",
       subtitle = "Trees range expanded to 200-500 based on diagnostic plot") +
  theme_minimal()

rf_fit |> show_best(metric = "roc_auc")

best_rf      <- rf_fit |> select_best(metric = "roc_auc")
final_rf_wf  <- rf_wf  |> finalize_workflow(best_rf)
final_rf_fit <- final_rf_wf |> last_fit(sepsis_split)

final_rf_fit |> collect_metrics()

final_rf_fit |>
  extract_fit_parsnip() |>
  vip(num_features = 19) +
  labs(title = "Random Forest Variable Importance")

final_rf_fit |>
  collect_predictions() |>
  roc_curve(prolonged_stay, .pred_1) |>
  autoplot()

rf_auc <- final_rf_fit |>
  collect_metrics() |>
  filter(.metric == "roc_auc") |>
  pull(.estimate)

cat("Random Forest Test AUC:", round(rf_auc, 3), "\n")

```

# XG Boost Model

```{r}
# ── XGBoost ───────────────────────────────────────────────────────────────────
gb_recipe <- recipe(prolonged_stay ~ ., data = sepsis_other) |>
  step_dummy(all_nominal_predictors()) |>
  step_zv(all_numeric_predictors())

gb_mod <- boost_tree(
  mode       = "classification",
  trees      = 1000,
  tree_depth = tune(),
  learn_rate = tune()
) |>
  set_engine("xgboost")

gb_wf <- workflow() |>
  add_recipe(gb_recipe) |>
  add_model(gb_mod)

# updated: learn_rate narrowed from (-5, 2) to (-5, 0)
# diagnostic plot showed AUC crashes after learn_rate > 1
# concentrates all grid points in the effective region (1e-05 to 1)
param_grid_gb <- grid_regular(
  tree_depth(range = c(1L, 3L)),
  learn_rate(range = c(-5, 0), trans = log10_trans()),
  levels = c(3, 10)
)

set.seed(42)
folds <- vfold_cv(sepsis_other, v = 5, strata = "prolonged_stay")

gb_fit <- gb_wf |>
  tune_grid(
    resamples = folds,
    grid      = param_grid_gb,
    metrics   = metric_set(roc_auc, accuracy)
  )

# diagnosis plot
gb_fit |>
  collect_metrics() |>
  filter(.metric == "roc_auc") |>
  mutate(tree_depth = as.factor(tree_depth)) |>
  ggplot(aes(x = learn_rate, y = mean, color = tree_depth)) +
  geom_point() + geom_line() +
  scale_x_log10() +
  geom_errorbar(aes(ymin=mean-std_err, ymax=mean+std_err), width=0.1) +
  labs(x = "Learning Rate (log scale)", y = "CV AUC",
       title = "XGBoost CV Tuning",
       subtitle = "Learn rate narrowed to 1e-05 to 1 based on diagnostic plot") +
  theme_minimal()

gb_fit |> show_best(metric = "roc_auc")

best_gb      <- gb_fit |> select_best(metric = "roc_auc")
final_gb_wf  <- gb_wf  |> finalize_workflow(best_gb)
final_gb_fit <- final_gb_wf |> last_fit(sepsis_split)

final_gb_fit |> collect_metrics()

final_gb_fit |>
  extract_fit_parsnip() |>
  vip(num_features = 19) +
  labs(title = "XGBoost Variable Importance")

final_gb_fit |>
  collect_predictions() |>
  roc_curve(prolonged_stay, .pred_1) |>
  autoplot()

gb_auc <- final_gb_fit |>
  collect_metrics() |>
  filter(.metric == "roc_auc") |>
  pull(.estimate)

cat("XGBoost Test AUC:", round(gb_auc, 3), "\n")
```

1.  Cohort flow diagram

```{r}
# cohort flow - how many patients at each step
cohort_flow <- tibble(
  Step = c(
    "Adult sepsis ICU admissions",
    "After outcome filter (LOS > 0)",
    "After first admission filter",
    "After 48h early death exclusion",
    "Final analytic cohort"
  ),
  N = c(5017, 4953, 4478, 4179, 4179),
  label = c(
    "5,017",
    "4,953 (-64 same-day discharges)",
    "4,478 (-475 duplicate admissions)",
    "4,179 (-299 early deaths)",
    "4,179 (final)"
  )
) |>
  mutate(Step = factor(Step, levels = rev(Step)))

ggplot(cohort_flow, aes(x = N, y = Step)) +
  geom_col(fill = "steelblue", width = 0.5) +
  geom_text(aes(label = label), hjust = -0.05, size = 3.5) +
  labs(title = "Cohort Construction Flow",
       x = "Number of Patients", y = NULL) +
  xlim(0, 9000) +   # increased to fit long labels
  theme_minimal()
```

2.  Key predictor distributions by outcome:

```{r}
# key continuous predictors based on LASSO + clinical importance
sepsis_imputed |>
  pivot_longer(
    cols = c(gcs_total, creatinine, lactate, 
             temperature, map, anchor_age),
    names_to  = "variable",
    values_to = "value"
  ) |>
  mutate(
    prolonged_stay = factor(prolonged_stay,
                            labels = c("Normal (<14d)", "Prolonged (≥14d)")),
    variable = recode(variable,
      gcs_total   = "GCS Total",
      creatinine  = "Creatinine",
      lactate     = "Lactate",
      temperature = "Temperature",
      map         = "MAP",
      anchor_age  = "Age"
    )
  ) |>
  ggplot(aes(x = value, fill = prolonged_stay)) +
  geom_density(alpha = 0.5) +
  facet_wrap(~ variable, scales = "free", ncol = 3) +
  scale_fill_manual(values = c("aquamarine3", "salmon")) +
  labs(title = "Distribution of Key Continuous Predictors by Outcome",
       x = NULL, y = "Density", fill = "Outcome") +
  theme_minimal() +
  theme(legend.position = "bottom")
```

3.  Comorbidity prevalence by outcome:

```{r}
sepsis_imputed |>
  mutate(prolonged_stay = factor(prolonged_stay,
                                  labels = c("Normal", "Prolonged"))) |>
  group_by(prolonged_stay) |>
  summarise(across(c(hypertension, diabetes, ckd, cvd,
                     chronic_liver, immunosuppression, vasopressor),
                   ~ round(mean(.x) * 100, 1))) |>
  pivot_longer(-prolonged_stay, 
               names_to = "comorbidity", 
               values_to = "pct") |>
  mutate(comorbidity = recode(comorbidity,
    hypertension      = "Hypertension",
    diabetes          = "Diabetes",
    ckd               = "CKD",
    cvd               = "CVD",
    chronic_liver     = "Chronic Liver",
    immunosuppression = "Immunosuppression",
    vasopressor       = "Vasopressor"
  )) |>
  ggplot(aes(x = reorder(comorbidity, pct), y = pct, fill = prolonged_stay)) +
  geom_col(position = "dodge", width = 0.6) +
  coord_flip() +
  scale_fill_manual(values = c("aquamarine3", "salmon")) +
  labs(title = "Comorbidity Prevalence by Outcome",
       x = NULL, y = "Prevalence (%)", fill = "Outcome") +
  theme_minimal() +
  theme(legend.position = "bottom")
```

```{r}
# RF Diagnosis Plot — confirms trees range was appropriate
rf_fit |>
  collect_metrics() |>
  filter(.metric == "roc_auc") |>
  mutate(mtry = as.factor(mtry)) |>
  ggplot(aes(x=trees, y=mean, color=mtry, group=mtry)) +
  geom_line() + geom_point() +
  geom_errorbar(aes(ymin=mean-std_err, ymax=mean+std_err), width=0.2) +
  labs(title="Random Forest CV Tuning — Hyperparameter Diagnostic",
       subtitle="Best mtry=3 at trees=400, interior to grid — range was appropriate",
       x="Number of Trees", y="CV AUC") +
  theme_minimal()

# XGBoost Diagnosis Plot — confirmed learn_rate range needed narrowing
gb_fit |>
  collect_metrics() |>
  filter(.metric == "roc_auc") |>
  mutate(tree_depth = as.factor(tree_depth)) |>
  ggplot(aes(x=learn_rate, y=mean, color=tree_depth)) +
  geom_point() + geom_line() +
  scale_x_log10() +
  geom_errorbar(aes(ymin=mean-std_err, ymax=mean+std_err), width=0.1) +
  labs(title="XGBoost CV Tuning — Hyperparameter Diagnostic",
       subtitle="Peak at learn_rate~0.006, interior to narrowed range — grid was appropriate",
       x="Learning Rate (log scale)", y="CV AUC") +
  theme_minimal()
```

```{r}
app_data <- list(
  cohort      = sepsis_imputed,
  lasso_model = lasso_model
)

saveRDS(app_data, "/Users/charlottelee/Documents/biostat-203b-2026-winter/final_project/app_data.rds")
cat("Saved! File size:", 
    round(file.size("/Users/charlottelee/Documents/biostat-203b-2026-winter/final_project/app_data.rds")/1e6, 1), 
    "MB\n")
```
